### 2025-05-07 writing/manuscript/main.tex
Moves to docker

% For styling
\documentclass{article}
\setlength\parindent{0pt}
\setlength{\parskip}{\baselineskip}
\usepackage[hidelinks]{hyperref}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{a4wide}
\usepackage{textgreek}

% For pandas dataframe
\usepackage{booktabs}
\usepackage{longtable}

% For referencing
%\let\cite\shortcite
%\bibliographystyle{apacite}
\bibliographystyle{plain}
%\bibliographystyle{ksfh_nat}
%%\usepackage[super]{natbib}
%%\usepackage{hyperref}
%%\bibliographystyle{unsrtnat}

%\usepackage[backend=bibtex,
%  style=nature,
%  sorting=none,
%  giveninits=true,
%  isbn=false,doi=false,url=false,
%  natbib=true
%]{biblatex}
%\addbibresource{library.bib}
%\usepackage{hyperref}

% For glossaries
\usepackage[nonumberlist]{glossaries}
\makeglossaries
\newacronym{SHAP}{SHAP}{SHapley Additive exPlanations}
\newacronym{MAM}{MAM}{Moderate Acute Malnutrition}
\newacronym{EC}{EC}{Expressive Communication}
\newacronym{EEG}{EEG}{Electroencephalography}
\newacronym{AUCROC}{AUCROC}{Area Under the Curve}
\newacronym{LMIC}{LMIC}{Low- and Middle- Income Countries}
\newacronym{MWU}{MWU}{Mann-Whitney U test}
\newacronym{WLZ/WHZ}{WLZ/WHZ}{weight-for-length/height}
\newacronym{PSD}{PSD}{power spectral density}
\newacronym{BBB}{BBB}{Blood Brain Barrier}
\newacronym{MUAC}{MUAC}{Mid-upper arm circumference}
\newacronym{PCoA}{PCoA}{Principal Coordinates Analysis}
\newacronym{PERMANOVA}{PERMANOVA}{PERmutational Multivariate ANalysis Of VAriance}
\newacronym{FDR}{FDR}{False Discovery Rate}
\newacronym{HC}{HC}{Head Circumference}
\newacronym{SCFA}{SCFA}{Short Chain Fatty Acid}
\newacronym{P/B}{P/B}{\textit{Prevotella}-to-\textit{Bacteroides}}
\newacronym{OCFA}{OCFA}{Odd Chain Fatty Acid}

% For drafts
\usepackage{setspace}
\doublespacing
\usepackage{lineno}
\linenumbers

% For captions at end of document
%\usepackage{figcaps}

\title{Linking the Gut Microbiome to Neurocognitive Development in Bangladesh Malnourished Infants}
\author[1,*]{Theo Portlock}
\author[2,*]{Talat Sharma}
\author[2,*]{Shahria Hafiz Kakon}
\author[3,*]{Berit Hartjen}
\author[1,*]{Chris Pook}
\author[1,*]{Brooke Wilson}
\author[3]{Ayisha Bhuttor}
\author[1]{Daniel Ho}
\author[1]{Inoli Shennon Wadumesthrige Don}
\author[3]{Anne-Michelle Engelstad}
\author[3]{Renata Di Lorenzo}
\author[3]{Garrett Greaves}
\author[3]{Caroline Kelsey}
\author[1]{Peter Gluckman}
\author[1,4,5,6]{Justin O'Sullivan}
\author[7]{Terrence Forrester}
\author[3]{Charles Nelson}

\affil[1]{The Liggins Institute, University of Auckland, NZ}
\affil[2]{Infectious Diseases Division, International Centre for Diarrheal Disease Research, Bangladesh}
\affil[3]{Department of Pediatrics, Boston Children’s Hospital and Harvard Medical School; Harvard Graduate School of Education, Boston, USA}
\affil[4]{The Maurice Wilkins Centre, The University of Auckland, New Zealand}
\affil[5]{MRC Lifecourse Epidemiology Unit, University of Southampton, University Road, Southampton, UK}
\affil[6]{Singapore Institute for Clinical Sciences, Agency for Science Technology and Research, Singapore}
\affil[7]{Faculty of Medical Sciences, UWI Solutions for Developing Countries, The University of the West Indies (UWI), Jamaica}
\affil[*]{These authors contributed equally}
\date{\vspace{-5ex}}

\begin{document}

\maketitle
\newpage
\printglossaries

\section*{Abstract}

\section*{Main}

\section*{Results}
\subsection*{Study population characteristics}
As a city with the second highest density of population and in a country with childhood malnutrition rate is one of the highest globally, the Mirpur region in Dhaka, Bangladesh was chosen to assess the impact of early-life malnutrition \cite{ahmed2012nutrition}.

\begin{figure}[!htb]
\centering
\includegraphics[scale=0.9]{../../figures/Figure1-microbiome.pdf}
\caption[Malnutrition impacts the 12-month-old infant gut microbiome]{
	Malnutrition impacts the 12-month-old infant gut microbiome.
	a) Schematic of study design.
	b) Summary of data collected.
	c) Change in diversity of the gut microbiome associated with malnutrition.
	d) PCoA Scatterplot of Bray-Curtis beta diversities of samples (each marker is a single infant sample).
	e) Barplot of significant taxonomic differences in relative abundance and prevailance between 12-month-old well-nourished and \gls{MAM} samples (\textit{p} \textless{} 0.05).
	f) Boxplot of \gls{P/B} ratio change between study conditions.
	g) Volcano plot of pathways that associate with malnutrition (red and orange horizontal line signifies \textit{q} \textless{} 0.05 and 0.01 respectively. left and right vertical lines represent Log\textsubscript{2}(\gls{MAM}/well-nourished) of -0.1 and 0.1 respectively).}
\label{Figure1}
\end{figure}

\section*{Discussion}

\section*{Conclusion}

\section*{Methods}
\subsection*{Ethics}
The M4EFaD intervention was registered NCT05629624 on clinicaltrials.gov.
The study was approved by icddr,b Ethical Review Committee PR-21084 and the Bangladesh Directorate General of Drug Administration.
Ethical review for the analytical component was obtained from Auckland Health Research Ethics Committee approval AH23922 (metabolomics, metagenomics, machine learning).

\subsection*{Study Design and Participants}
The study was performed on the baseline data from three cohorts of infants who were enrolled (between Jan – December 2022) as part of the M4EFaD intervention within the Mirpur slum, Dhaka, Bangladesh.
Inclusion criteria included a diagnosis of malnutrition, no history of chronic medical conditions, and no antibiotic use within the past month.

\subsection*{Recruitment and anthropometric data collection}
Enrolment was initiated on February 7, 2022, and will continue until February 2024.
Study surveillance workers (SWs) conducted a door-to-door census (approximately 100,000 households) in Mirpur DNCC wards ward 2, 3 and 5 between January and December 2022.
Verbal consent was obtained to participate in the census.
The census identified 5736 children aged between 11 to 13 months and 2,314 children aged between 34 to 38 months.
During the census, if the guardian verbally consented to the study procedure, and the babies met the inclusion and exclusion criteria of the study (Table 1), the SWs proceeded to measure the \gls{MUAC} of the child.
Mothers of babies who were within the \gls{MUAC} range were invited to visit the icddr,b study clinic for further assessment and enrolment.
Final screening for eligibility and study consent occurred at the icddr,b Mirpur study clinic.
The consenting process was tailored to each mother's literacy level and involved reviewing the inclusion and exclusion criteria.
Comprehension of the study was assessed using scripted points and open-ended questions.
Following consent, the clinical screening team completed a screening form, capturing the date of enrolment, sex, date of birth (DOB), weight (in kg), length/ height (in cm), head circumference (in cm), and Mid-Upper Arm Circumference \gls{MUAC} measurements of the child.
The \gls{WLZ/WHZ} Z-score for each child was calculated using the WHO anthropometric calculator.
The child's age was validated using the EPI vaccination card.
Neurological measures, Bailey scores, EEG data were collected upon enrolment to evaluate neurological development.

\subsection*{EEG data collection and analysis}
Continuous scalp EEG was recorded using NetStation 4.5.4. and 128-channel Hydrocel Geodesic Sensor Nets modified to remove eye electrodes (Electrical Geodesics, Inc. (EGI), Eugene, OR, USA).
Data was sampled at 500 Hz.
Impedances were kept under 100 k \textomega{} when possible and measured once at the beginning of the session, and again halfway through.
Sessions were conducted in a dimly lit room with the participants sitting on the parent’s lap.
The participants were separated from the research staff conducting the session by a curtain, but the testing area was not acoustically or electrically shielded.
A second research staff member was present in the testing area to help keep the participant engaged.
EEG sessions consisted of 6 paradigms, i.e., resting state, visual working memory, flanker, disengagement, visual evoked potential, and auditory stimuli.
The subsequent (pre-)processing steps were applied to the resting state data where participants watched a 3-minute video that featured toys.

EEG data were preprocessed offline with MatLab (R2021B) using the Harvard Automated Processing Pipeline for Electroencephalography (HAPPE) Version 3 (Gabard-Durnam et al., 2018).
A specified subset of 30 channels was excluded (‘E1’, ’E8’, ’E14’, ’E17’, ’E21’, ‘E25’, ’E32’, ‘E38’, ‘E43’, ’E44’,’ E48’, ’E49’, ’E56’, ’E63’, ’E68’, ’E73’, ’E81’, ’E88’, ’E94’, ’E99’, ’E107’, ’E113’, ’E114’, ’E119’, ’E120’, ’E121’, ‘E125', 'E126', 'E127', 'E128').
Data were downsampled to 250Hz, bandpass filtered (1-100Hz), and filtered using a 50Hz cleanline filter for line noise removal.
Bad channels were then automatically identified and rejected, and wavelet-enhanced Independent Component Analysis (ICA) and the Multiple Artifact Rejection Algorithm (MARA) were performed to detect and impute artifacts.
Resting state data were segmented into 2s epochs; epochs with an amplitude >±150mV were rejected.
Segments were also rejected using segment similarity criteria.
Data were then re-referenced to the average of all channels.

EEG outputs from HAPPE were then reformatted and processed using the Batch Electroencephalography Automated Processing Platform (BEAPP) (Levin et al., 2018) to extract power spectra for each participant across the following frequency bands: delta (2-4Hz), theta (4-6Hz), low alpha (6-9Hz), high alpha (9-12Hz), beta (12-30Hz), and gamma (30-45Hz) and the following regions of interest (see Supp Figure 2): occipital (‘E70’, ’E71’, ’E75’, ‘E76’, ‘E83’), temporal (‘E36’,‘E40’, ‘E41’, ‘E45’, ‘E46’, ‘E102’, ‘E103’, ‘E104’, ‘E108’, ‘E109’), parietal (‘E52’, ‘E53’, ‘E59’, ‘E60’, ‘E85’, ‘E86’, ‘E91’, ‘E92’), and frontal (‘E5’, ‘E6’, ‘E12’, ‘E13’, ‘E24’, ‘E27’, ‘E28’, ‘E33’, ‘E34’, ‘E112’, ‘E116’, ‘E117’, ‘E122’, ‘E123’, ‘E124’).
Further, PSD values were normalized by a Log\textsubscript{10} transform.

\subsection*{Developmental Outcomes} 
The Bayley Scales of Infant and Toddler Development, Fourth Edition (BSID-IV) cognitive, language, and motor subscales were administered to all participants.
Research assistants were trained to research reliability in the administration and scoring of the Bayley-4.
Due to cultural differences between the Bangladesh and the United States where the assessment was developed, Bangladeshi researchers modified some assessment stimuli to improve cultural responsiveness and relevancy.
For example, pictures for the item naming series and action naming series of the expressive language and receptive language subscales were adapted to include items that Bangladeshi children are more likely to be familiar with and bedtime clothing that would signify the child in the picture was going to sleep instead of the one-piece pajamas worn in the original picture, which the Bangladeshi children would not be familiar with.

\subsection*{Biological sample collection}
Stool samples were collected from each infant at their home at the baseline visit.
Samples were collected in DNA/RNA Shield Fecal Collection Tubes (Zymo Research, \#R1101).
Peripheral venous blood samples were collected in EDTA Vacutainers, separated into plasma and RBCs and immediately frozen at -80 C.
Batches of blood and stool samples were air-freighted on dry ice from Bangladesh to the Liggins Institute, New Zealand for processing and analysis.

\subsection*{Microbiome DNA extraction and sequencing}
DNA was extracted from stool samples using the ZymoBIOMICS MagBead DNA/RNA extraction kit (Zymo Research, \#R2136) following the standard protocol.
Samples (1 mL) were mechanically lysed in bead bashing tubes using the MiniG tissue homogenizer prior to extraction of DNA.
200 µL of the sample was used post-bead bashing for extraction of DNA following the protocol.
A volume of 50 \textmu{} L of elute was collected in DNAse/RNAse Free Water.
Samples with a DNA concentration \textless{} 14.5 ng/\textmu{}L were re-extracted following the ZymoBIOMICS DNA extraction protocol.
Samples were sequenced (Illumina NovaSeq 150PE reads) to an average sequencing depth of 20M read-pairs/sample.
Raw sequences were processed using BioBakery3 tools \cite{beghini2021integrating}, specifically read quality filtering and human decontamination with KneadData (Version 1), taxonomic profiling with MetaPhlAn3 (Version 3.1, using the mpa\_v31\_CHOCOPhlAn\_201901 database) and functional profiling using presence/absence and abundance of microbial pathways (MetaCyc) with HUMAnN3 (Version 3.6).
A minimum threshold of \textgreater{} 0.1\% relative abundance and \textgreater{} 5\% prevalence for all detected species was applied.

\subsection*{Plasma lipidomics}
Plasma samples for lipidomics were thawed on ice and extracted according to a method modified from \citet{liu2016plasma}.
Briefly, 10 \textmu{}L volume was placed in an amber glass autosampler vial and 300 \textmu{}L of a mixture of Type 1 water, butanol, methanol, chloroform and SPLASH Lipidomix in a ratio of 4:15:15:20:1 was added.
The mixture was vortexed and sonicated at room temperature before the protein precipitate was removed by centrifugation and an aliquot of supernatant transferred to an amber glass autosampler for negative ionisation LC-MS/MS.
A second aliquot of supernatant was diluted 5 times with 75\% IPA for positive ionisation LC-MS/MS.
A 5 µL volume of each sample was injected onto a Phenomenex Kinetex F5 column (100 mm × 2.1 mm × 2.6 \textmu{}m) and lipids were separated using a ternary gradient of Type 1 water, methanol and isopropanol containing ammonium acetate.
Lipids were quantified and identified with a Q-Exactive mass spectrometer (Thermo Fisher Scientific, Germany) equipped with a heated electrospray ionisation HESI source.
Data was processed using MS-DIAL v4.92 92 \cite{tsugawa2015ms}.
For full methodological details see the supplementary information.

\subsection*{Statistical Analyses}
Python version 3.9.2 was used to perform all analysis \cite{van1995python}.
Due to the unequal sample sizes and non-normally distributed data; non-parametric statistical approaches were used for differential abundance analysis.
Relative abundances were adjusted by Centred Log Ratio to account for the compositional nature of the dataset \cite{gloor2016s}.
Log adjusted fold change significance was measured using (\gls{MWU}) test using the ‘mannwhitneyu’ function from ‘scipy.stats’ and adjusted for multiple testing using the ‘fdrcorrection’ function from statsmodels.stats.multitest.
\gls{PCoA} ordinations (plotted using 'skbio.stats.ordination.pcoa' module) were used to visualise the clustering of the Bray-Curtis dissimilarities (calculated using skbio.distance.pdist) between samples from their species and functional composition.
To quantify the variance of the gut microbiome explained covariates, \gls{PERMANOVA} p-values were calculated from those Bray-Curtis Dissimilarities using the ’permanova’ function from the 'skbio.stats.distance' module.
Bray-Curtis were also used to capture the temporal dynamics of the microbiome from baseline.
Numerical Associations between species and metadata were measured with Spearman correlation (calculated using 'spearmanr' function from 'scipy.stats' module), where significance was defined as \gls{FDR} adjusted p-values of \textless{} 0.05 as per \citet{2020SciPyNMeth}.
Associations between categorical data were measured with Fisher's Exact test (calculated using 'fisher\_exact' from 'scipy.stats' module), where significance was defined as p-values of \textless{} 0.05.

\subsection*{Machine learning}
SHAP Value (SHapley Additive exPlanations) interpretation was used to interpret the contributions each feature had on the model's performance using the ‘shap’ python package \cite{lundberg2017unified}.

\subsection*{Network analysis}
Absolute spearman rho of above 0.3 were used as edges and gut bacterial species and functional profiles, EEG, and plasma lipids were used as nodes coloured by their mean SHAP scores for classifier models that distinguish \gls{MAM} from well-nourished conditions.

\section*{Code availability}
All analysis code is available on the GitHub repository.
The codebase is organised into scripts, providing a comprehensive framework for replicating the experiments.
Detailed documentation and instructions on how to use the code are provided in the repository's README file.

\section*{Ethics approval and consent to participate}
Ethical approvals were obtained from the Research Review Committee (RRC; August 21, 2021) and Ethical Review Committee (ERC) of icddr,b (protocol no: PR-21084; September 21, 2021), Institutional Review Board of Boston Children’s Hospital, USA (for analyses of neuropsychological assessments), University of Auckland, New Zealand (approval AH23922; for analyses of collected biological samples) and University of West Indies (CREC-MN.51, 21/22).

\section*{Data availability}
EEG and metadata are available from the authors, upon reasonable request that meets the ethics of the study.

\section*{Competing interests}
The authors declare that they have no competing interests.

\section*{Funding}
Work on this clinical trial is supported by Wellcome Leap (9942 Culver Blvd Unit 1277 Culver City, CA 90232-4167, United States; www.wellcomeleap.org) to PDG, JMO, TF and CAN as part of the 1kD Program.
We acknowledge our core donors, Governments of Bangladesh, Canada for providing unrestricted support and commitment to icddr,b's research effort.

\section*{Author Contributions}
TP, KG and JOS drafted and co-wrote the manuscript.
TS, SHK, BCW, BH, CP, AB, DH, IS, AME, RD, GG, CK, PDG, RH, TF, CAN commented on the manuscript.
JMO, RH, TF, PDG, CAN designed the study and analyses.
TS, SHK performed assessments and obtained samples in Dhaka.
RH oversaw the Dhaka group.
TP performed multiomic analyses, BCW and IS performed metagenomics, CP performed metabolomics, JOS oversaw the Auckland group.
BH performed EEG analyses, CAN oversaw the Boston group.

\section*{Acknowledgements}
The authors would like to acknowledge the participants in Mirpur, Dhaka, Bangladesh for their contributions to this study.
The authors would also like to thank the study team within the Infectious Diseases Division, International Centre for Diarrheal Disease Research, Bangladesh for their work in participant recruitment, sample collection and assessments.

\bibliography{library}
\printbibliography

\section*{Supplementary material}

\begin{table}[!htb]
\centering
\caption[Baseline infant characteristics]{
	Baseline infant characteristics.
	Plus minus values are means \textpm{} SD from continous variables and their pvalues are calculated using \gls{MWU}.
	All other variables are categorical (True vs False) with their pvalues calculated using Fishers Exact test.}
%\input{../../figures/texsupptables/TableS1_patientinfo}
\label{TableS1}
\end{table}

\end{document}

### 2025-05-07 code/anthrotime.py

@@ -10,9 +10,14 @@ df = anthro.join(timemeta)
 
 # Add condition to timemeta
 df.loc[df.Condition == 'Well-nourished', 'Recovery'] = 'Healthy'
+df.loc[df.Condition == 'Well-nourished', 'Feed'] = 'Healthy'
 
 # Define custom color palette
-palette = {'Healthy':'magenta','Recovered': 'orange', 'No recovery': 'red'}
+#palette = {'Healthy':'magenta','Recovered': 'orange', 'No recovery': 'red'}
+palette = {'Healthy':'magenta','ERUSF (B)': 'blue', 'Local RUSF (A)': 'red'}
+
+# Drop all after 52 weeks
+df = df.loc[df.timepoint <= 52]
 
 # Function to plot individual lines and mean trend
 def line(df, x, y, hue, ax):
@@ -40,9 +45,9 @@ def line(df, x, y, hue, ax):
 fig, axs = plt.subplots(1, 3, figsize=(6, 2.5), sharex=True)
 
 # Plot each metric
-line(df, 'timepoint', 'WLZ_WHZ', 'Recovery', axs[0])
-line(df, 'timepoint', 'Length', 'Recovery', axs[1])
-line(df, 'timepoint', 'Weight', 'Recovery', axs[2])
+line(df, 'timepoint', 'WLZ_WHZ', 'Feed', axs[0])
+line(df, 'timepoint', 'MUAC', 'Feed', axs[1])
+line(df, 'timepoint', 'Weight', 'Feed', axs[2])
 
 # Layout tweaks
 plt.subplots_adjust(wspace=0.4)

### 2025-05-07 code/calculate_diversity.sh

@@ -4,9 +4,6 @@
 metaphlan_file=$1
 mkdir -p ../results/mbiome_alpha
 
-# Transpose taxo file
-transpose.py $metaphlan_file
-
 # For alpha diversity measures
 for metric in gini shannon richness simpson
 do

### 2025-05-07 code/create_mapping.py

@@ -1,25 +1,53 @@
+#!/usr/bin/env python
+
+import argparse
 import pandas as pd
-import sys
+from pathlib import Path
 
-# Use default file if none provided
-datatable = sys.argv[1] if len(sys.argv) > 1 else '../conf/timedatasets.txt'
+def parse_arguments():
+    """Parse command-line arguments."""
+    parser = argparse.ArgumentParser(description="Create subjectID/timepoint mapping from datasets.")
+    parser.add_argument('-i', '--input-file', type=Path, default=Path('../conf/timedatasets.txt'),
+                        help='Path to input file listing datasets (default: ../conf/timedatasets.txt)')
+    parser.add_argument('-d', '--datasets-dir', type=Path, default=Path('../results'),
+                        help='Directory where dataset files are located (default: ../results)')
+    parser.add_argument('-o', '--output-file', type=Path, default=Path('../results/mapping.tsv'),
+                        help='Path to output mapping TSV file (default: ../results/mapping.tsv)')
+    return parser.parse_args()
 
-# Read all dataset filenames listed in the datatable
-with open(datatable, 'r') as f:
-    lines = [line.strip() for line in f if line.strip()]
+def main():
+    args = parse_arguments()
 
-# Collect mappings from all files
-all_mappings = []
+    # Read dataset list
+    with args.input_file.open('r') as f:
+        lines = [line.strip() for line in f if line.strip()]
 
-for filepath in lines:
-    df = pd.read_csv(f'../results/{filepath}noyr3.tsv', sep='\t', index_col=0)
-    mapping = df.index.to_series().str.split('_', expand=True).iloc[:, :2]
-    mapping.columns = ['subjectID', 'timepoint']
-    mapping.index = df.index
-    all_mappings.append(mapping)
+    # Collect mappings
+    all_mappings = []
+    for filepath in lines:
+        dataset_path = args.datasets_dir / f'{filepath}.tsv'
+        if not dataset_path.exists():
+            print(f"Warning: {dataset_path} not found. Skipping.")
+            continue
+        df = pd.read_csv(dataset_path, sep='\t', index_col=0)
+        mapping = df.index.to_series().str.split('_', expand=True).iloc[:, :2]
+        mapping.columns = ['subjectID', 'timepoint']
+        mapping.index = df.index
+        all_mappings.append(mapping)
 
-# Merge all mappings into a single DataFrame
[…]

### 2025-05-07 code/create_table1.py

@@ -1,0 +1,56 @@
+#!/usr/bin/env python
+
+import pandas as pd
+
+df = pd.read_csv('../results/filtered/surveillance.tsv', sep='\t', index_col=0)
+meta = pd.read_csv('../results/filtered/meta.tsv', sep='\t', index_col=0)
+
+
+
+#!/usr/bin/env python
+
+import pandas as pd
+import seaborn as sns
+import matplotlib.pyplot as plt
+from scipy.stats import kruskal
+
+# Load data
+df = pd.read_csv('../results/filtered/surveillance.tsv', sep='\t', index_col=0)
+meta = pd.read_csv('../results/filtered/meta.tsv', sep='\t', index_col=0)
+
+# Merge on subjectID
+merged = df[['days_to_recovery']].join(meta)
+
+# Drop rows with missing days_to_recovery
+merged = merged.dropna(subset=['days_to_recovery'])
+
+# Define categorical columns to compare
+categorical_cols = ['Sex', 'Ethnicity', 'Condition', 'Feed', 'Recovery', 'Delivery_Mode', 'Supplementation', 'PoB']
+
+# Loop through each categorical variable
+for col in categorical_cols:
+    if col not in merged.columns:
+        continue
+
+    print(f"\n=== {col} ===")
+
+    # Grouped data
+    grouped = merged.groupby(col)['days_to_recovery'].apply(list)
+
+    # Ensure there are at least 2 groups with data
+    if len(grouped) < 2:
+        print("Not enough groups to compare.")
+        continue
+
+    # Statistical test
+    stat, p = kruskal(*grouped)
+    print(f"Kruskal-Wallis H-test: H={stat:.2f}, p={p:.4f}")
+
+    # Boxplot
+    plt.figure(figsize=(8, 5))
+    sns.boxplot(data=merged, x=col, y='days_to_recovery')
+    plt.title(f"Days to Recovery by {col}\nKruskal-Wallis p={p:.4f}")
+    plt.xticks(rotation=45)
+    plt.tight_layout()
+    plt.show()
+

### 2025-05-07 code/deltas.py

@@ -1,30 +1,52 @@
 #!/usr/bin/env python
+
+import argparse
+from pathlib import Path
 import pandas as pd
 import numpy as np
 
-# Function to compute log differences between baseline (week 0) and one-year (week 52)
-def calculatedeltas(filepath):
+def parse_arguments():
+    """Parse command-line arguments."""
+    parser = argparse.ArgumentParser(description="Compute log differences between baseline and one-year samples.")
+    parser.add_argument('-i', '--input', type=Path, required=True,
+                        help='Path to input dataset file (e.g., ../results/aanoyr3.tsv)')
+    parser.add_argument('-m', '--meta', type=Path, required=True,
+                        help='Path to timemeta.tsv file')
+    parser.add_argument('-o', '--output', type=Path, required=True,
+                        help='Path to output file (e.g., ../results/aadiff.tsv)')
+    return parser.parse_args()
+
+def calculate_deltas(df):
+    """Compute differences between baseline (week 0) and one-year (week 52)."""
     bl = df.query('timepoint == 0').drop(columns=['timepoint'])
     oneyr = df.query('timepoint == 52').drop(columns=['timepoint'])
     
-    # Ensure indices match before division
+    # Ensure indices match
     common_idx = bl.index.intersection(oneyr.index)
     bl, oneyr = bl.loc[common_idx], oneyr.loc[common_idx]
     
-    #diff = oneyr.div(bl).apply(np.log)  # Log fold-change
-    diff = oneyr.sub(bl)  # Just subtraction
+    diff = oneyr.sub(bl)
     return diff
 
-# Process timepoint datasets
-with open('../conf/timedatasets.txt') as f:
-    for filepath in f:
-        dataset_path = filepath.strip()
-        try:
-            dataset = pd.read_csv(f'../results/{dataset_path}noyr3.tsv', sep='\t', index_col=0)
-            timemeta = pd.read_csv(f'../results/timemeta.tsv', sep='\t', index_col=0)
-            df = timemeta[['subjectID','timepoint']].join(dataset, how='inner').set_index('subjectID')
-            diff = calculatedeltas(df)
-            diff.to_csv(f'../results/{dataset_path}diff.tsv', sep='\t')
-        except Exception as e:
-            print(f"Error processing {dataset_path}: {e}")
+def main():
+    args = parse_arguments()
 
+    # Load data
[…]

### 2025-05-07 code/format_explain_var.py

@@ -1,0 +1,14 @@
+import pandas as pd
+
+dataset_lables = pd.read_csv('../conf/dataset_labels.tsv', sep='\t', index_col=0)
+df = pd.read_csv('../results/aa_perm.tsv', sep='\t', index_col=0)
+
+df.index = df.index.str.replace('.tsv','').str.replace('.*/','', regex=True)
+df = df.dropna()
+
+df.drop(columns=['Unnamed: 1'], inplace=True)
+
+df = df.join(dataset_lables[['name','class']]).set_index('name')
+
+df.to_csv('../results/aa_perm_formatted.tsv', sep='\t')
+

### 2025-05-07 code/format_ses.py

@@ -110,6 +110,9 @@ df = df.loc[:, ~df.columns.isin(mediacols)]
 economiccols = df.columns[df.columns.str.contains('taka')].tolist()
 ecdf = df.loc[:, economiccols]
 ecdf.columns = ecdf.columns.str.replace('_(taka)','').str.replace('(','').str.replace(')','').str.replace(',','_')
+
+# Drop the other columns as they are not commonly used
+ecdf = ecdf.loc[:, ecdf.columns[~ecdf.columns.str.contains('Other')]]
 ecdf.to_csv('../results/economics.tsv', sep='\t')
 df = df.loc[:, ~df.columns.isin(economiccols)]
 

### 2025-05-07 code/format_surveillance.py

@@ -18,11 +18,12 @@ df = pd.concat([df1, df2], axis=0)
 
 # Convert symptom responses to boolean cleanly
 cols = ['Cough', 'Rash', 'Antibio', 'ORS', 'Zinc']
-df_bool = (df[cols] - 2).mul(-1).astype(bool)
+df_bool = (df[cols] - 2).mul(-1).fillna(0).astype(bool)
 df[cols] = df_bool.astype('boolean')  # Nullable BooleanDtype for compatibility
 
 # Drop irrelevant columns
-df.drop(['Food', 'Date', 'Recover', 'Reason', 'Remarks'], axis=1, inplace=True)
+#df.drop(['Food', 'Date', 'Recover', 'Reason', 'Remarks'], axis=1, inplace=True)
+df.drop(['Date', 'Recover', 'Remarks'], axis=1, inplace=True)
 
 # Index renaming
 df.index = 'LCC' + df.index.astype(str)
@@ -45,8 +46,55 @@ df['timepoint'] = df['timepoint'].apply(np.ceil).astype(int)
 df['subjectID'] = df.index
 df = df.set_index(['subjectID', 'timepoint'])
 
-# LOOK TO CREATE SUMMARY STATS FROM THIS TABLE
-# May have to revisit this
+# Work out days to recovery
+days_of_catchup = df.reset_index().groupby(['Food','subjectID'])['Day'].last().xs(1)
+#days_of_catchup[days_of_catchup == 90.0] = np.nan
+days_of_catchup = days_of_catchup.rename('days_of_catchup')
+
+## quick and dirty analyis
+#meta= pd.read_csv('../results/meta.tsv', sep='\t', index_col=0)
+#meta['dtr'] = days_to_recovery
+#sns.kdeplot(data=meta,x='dtr', hue='Feed')
+
+# Base core body temperature
+base_body_temp = df.reset_index().groupby(['subjectID'])['Temp'].first()
+base_body_temp = base_body_temp.rename('base_temp')
+
+# Comorbidities
+## Sort out vomiting and diarrhoea so that it's just at least once per day
+df['DS'] = df['DS'] != 0.0
+df['Vomiting'] = df['Vomiting'] != 0.0
+avs = df.reset_index().groupby(['subjectID']).mean()
+avs = avs[['Fever','DS','Cough','Rash','Antibio','ORS','Zinc']].astype(float)
+avs = avs.set_axis(['fever','diarrhoea','cough','rash','antibiotic','ors','zinc'], axis=1)
+
+# Comp foods
[…]

### 2025-05-07 code/heatmap_alltimedata.py

@@ -38,7 +38,7 @@ def main():
     # Load data
     meta = pd.read_csv('../results/timemeta.tsv', sep='\t', index_col=0)
     dataset_names = load_datasets(args.datasets_file)
-    datasets = {data: pd.read_csv(f'../results/{data}noyr3.tsv', sep='\t', index_col=0) for data in dataset_names}
+    datasets = {data: pd.read_csv(f'../results/filtered/{data}.tsv', sep='\t', index_col=0) for data in dataset_names}
     
     df = pd.concat(datasets, axis=1)
     

### 2025-05-07 code/individual_recovery.py

@@ -1,46 +1,0 @@
-#!/usr/bin/env python
-import matplotlib.pyplot as plt
-import numpy as np
-import pandas as pd
-import sys 
-import metatoolkit.functions as f
-import seaborn as sns
-
-# Load data
-#subject = 'taxoDist'
-subject = sys.argv[1]
-df = f.load(subject)
-
-# Load metadata
-timemeta = f.load('fulltimemeta')
-timemeta['Refeed'] = timemeta['Feeds_by_Randomization'].str.replace('.*\(','', regex=True).str.replace(')','').str.replace('Healthy','H', regex=True)
-timemeta = timemeta[['ID', 'timepoint','Refeed']]
-
-# Stack data
-df.index = df.index.set_names('source')
-df.columns = df.columns.set_names('target')
-stacked = df.stack().to_frame('dist').reset_index()
-
-# Map the ID and timepoint
-stacked = stacked.merge(timemeta, left_on='source', right_index=True).rename(columns={'ID':'source_ID','timepoint':'source_timepoint', 'Refeed':'source_refeed'})
-stacked = stacked.merge(timemeta, left_on='target', right_index=True).rename(columns={'ID':'target_ID','timepoint':'target_timepoint', 'Refeed':'target_refeed'})
-
-# filter from baseline
-t1distances = stacked.query("source_timepoint == 0 and target_timepoint == 0 and target_refeed == 'H'")
-t1distances = t1distances[['source_ID','dist']].groupby('source_ID').mean().set_axis(['t1_dist'], axis=1)
-
-t2distances = stacked.query("source_timepoint == 52 and target_timepoint == 52 and target_refeed == 'H'")
-t2distances = t2distances[['source_ID','dist']].groupby('source_ID').mean().set_axis(['t2_dist'], axis=1)
-
-outdf = pd.concat([t1distances, t2distances], axis=1)
-foutdf = outdf.dropna()
-foutdf['improvement'] = np.log(foutdf.t1_dist.div(foutdf.t2_dist))
-foutdf['improvement'] = outdf.t1_dist.div(foutdf.t2_dist)
-
-sns.histplot(foutdf.improvement)
-f.savefig(subject + 'improvementhist')
-
-# 25% improvement
-odf = foutdf.improvement.gt(1.25).to_frame(subject)
-f.save(odf, subject + 'recovered')
-

### 2025-05-07 code/new_plot_bar.py

@@ -1,0 +1,95 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+import argparse
+import os
+import pandas as pd
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+def parse_args():
+    parser = argparse.ArgumentParser(description='Horizontal bar plot from coefficients (generalized).')
+    parser.add_argument('input', help='Input TSV/CSV file')
+    parser.add_argument('-o', '--output', required=True, help='Output path for the SVG file')
+    parser.add_argument('--coef-col', required=True, help='Name of the column containing coefficient values')
+    parser.add_argument('--hue-col', help='Optional: Name of the column to color bars by (e.g., filename)')
+    parser.add_argument('--index-col', help='Optional: Name of the column to use as row labels (otherwise default index)')
+    parser.add_argument('--separator', default='\t', help='Field separator (default: tab)')
+    parser.add_argument('--xscale', default='linear', choices=['linear', 'log', 'symlog'], help='X-axis scale')
+    parser.add_argument('--xlabel', default='Explained variance (R2)', help='Label for the x-axis')
+    parser.add_argument('--ylabel', default='Entry', help='Label for the y-axis')
+    parser.add_argument('--title', default='Horizontal Barplot', help='Plot title')
+    return parser.parse_args()
+
+def plot_horizontal_bar(df, coef_col, hue_col, output_path, xscale, xlabel, ylabel, title):
+    # Sort by coefficient column
+    df = df.sort_values(coef_col, ascending=False)
+
+    plt.figure(figsize=(5, max(4, len(df) * 0.15)))  # dynamic height
+
+    # Set up barplot parameters
+    barplot_kwargs = {
+        'x': coef_col,
+        'y': df.index,
+        'data': df,
+        'dodge': False,
+        'orient': 'h'
+    }
+
+    if hue_col:
+        barplot_kwargs['hue'] = hue_col
+        barplot_kwargs['palette'] = 'tab10'
+
+    ax = sns.barplot(**barplot_kwargs)
+
+    ax.set_xscale(xscale)
+    plt.xlabel(xlabel + (f" ({xscale} scale)" if xscale != 'linear' else ''))
+    plt.ylabel(ylabel)
+    plt.title(title)
+
+    # Handle legend
+    if hue_col:
+        plt.legend(title=hue_col, bbox_to_anchor=(1.05, 1), loc='upper left')
+    else:
[…]

### 2025-05-07 code/overall_recovery.py

@@ -6,11 +6,11 @@ import metatoolkit.functions as f
 # Load data
 #subject = 'anthro'
 subject = sys.argv[1]
+dataset = pd.read_csv(subject, sep='\t', index_col=0)
 dataset = f.load(subject)
-#dataset = dataset.drop('HC', axis=1).dropna()
 
 # Load metadata
-meta = f.load('timemeta')
+meta = pd.read_csv('../results/timemeta.tsv', sep='\t', index_col=0)
 meta['Refeed'] = meta['Feed'].fillna('Healthy').str.replace('.*\(','', regex=True).str.replace(')','').str.replace('Healthy','H', regex=True)
 
 # Calculate |Ct2-Mt2X| / |Ct1-Mt1X| 
@@ -49,7 +49,7 @@ y = 100*((t1y/t2y)-1)
 
 df = pd.concat([Ct1,Ct2,Mt1Y,Mt1X,Mt2Y,Mt2X])
 pcoa = f.calculate('pcoa', df)
-#f.setupplot()
+
 f.spindle(pcoa)
 f.savefig(subject+'recoveryspindle')
 
@@ -61,6 +61,6 @@ t1ysig = f.PERMANOVA(pd.concat([Ct1, Mt1Y]))
 output = pd.DataFrame([x, y, t1xsig, t1ysig, t2xsig,t2ysig], columns=[subject], index=['A_improvement(%)', 'B_improvement(%)', 'A_t1_pval', 'B_t1_pval', 'A_t2_pval', 'B_t2_pval']).T
 print(output)
 
-f.save(output, subject+'recovery')
+output.to_csv('../results/{subject}recovery.tsv', sep='\t')
 
 

### 2025-05-07 code/plot_bar.py

@@ -1,0 +1,58 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+import argparse
+import os
+import pandas as pd
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+def parse_args():
+    parser = argparse.ArgumentParser(description='Horizontal bar plot from coefficients.')
+    parser.add_argument('input', help='Input TSV file with columns coef and filename')
+    parser.add_argument('-o', '--output', required=True, help='Output path for the SVG file')
+    return parser.parse_args()
+
+def plot_horizontal_bar(df, output_path):
+    # Sort by coef
+    df = df.sort_values('coef', ascending=True)
+
+    # Set up the figure
+    plt.figure(figsize=(6, max(4, len(df) * 0.15)))  # dynamic height
+    ax = sns.barplot(
+        x='coef',
+        y=df.index,
+        hue='filename',
+        data=df,
+        dodge=False,
+        palette='tab10',
+        orient='h'
+    )
+
+    ax.set_xscale('symlog')  # <-- Symmetric log scale on x-axis
+
+    plt.xlabel('Coefficient (symlog scale)')
+    plt.ylabel('Entry')
+    plt.title('Horizontal Barplot of Coefficients by Filename')
+    plt.legend(title='Filename', bbox_to_anchor=(1.05, 1), loc='upper left')
+    plt.tight_layout()
+
+    # Make sure output directory exists
+    os.makedirs(os.path.dirname(output_path), exist_ok=True)
+
+    # Save the plot as SVG
+    plt.savefig(output_path, format='svg')
+    plt.close()
+    print(f"Plot saved to {output_path}")
+
+def main():
+    args = parse_args()
+    
+    # Load data
+    df = pd.read_csv(args.input, sep='\t', index_col=0)
+
+    # Call the plotting function
+    plot_horizontal_bar(df, args.output)
+
+if __name__ == '__main__':
+    main()

### 2025-05-07 code/plot_recover.py

@@ -18,7 +18,7 @@ f.clustermap(nndf, figsize=(1.2,4))
 f.savefig('recoverycluster')
 
 f.polar(ndf.T)
-plt.yscale('symlog')  # Set y-axis to symlog scale
+plt.yscale('symlog')
 f.savefig('polarrecovery')
 
 

### 2025-05-07 code/remove_3yr_control.py

@@ -1,19 +1,38 @@
 #!/usr/bin/env python
 
+import argparse
 import pandas as pd
-import sys
 from pathlib import Path
 
-# Read input
-input_file = Path(sys.argv[1])
-df = pd.read_csv(input_file, index_col=0, sep='\t')
+def parse_arguments():
+    """Parse command-line arguments."""
+    parser = argparse.ArgumentParser(description="Filter rows where 4th character in index is not '3'.")
+    parser.add_argument('input_file', type=Path, help='Path to input TSV file')
+    parser.add_argument('-o', '--output-file', type=Path, default=None, help='Optional path to save output file')
+    return parser.parse_args()
 
-# Filter
-df = df.loc[df.index.str[3] != '3']
+def main():
+    args = parse_arguments()
 
-# Prepare output path
-output_dir = input_file.parent
-output_file = output_dir / f"{input_file.stem}noyr3.tsv"
+    # Read input
+    df = pd.read_csv(args.input_file, index_col=0, sep='\t')
 
-# Save
-df.to_csv(output_file, sep="\t", index=True)
+    # Filter
+    df = df.loc[df.index.str[3] != '3']
+
+    # Prepare output path
+    if args.output_file is None:
+        output_dir = args.input_file.parent
+        output_file = output_dir / f"{args.input_file.stem}noyr3.tsv"
+    else:
+        output_file = args.output_file
+
+    # Ensure output directory exists
+    output_file.parent.mkdir(parents=True, exist_ok=True)
+
+    # Save
+    df.to_csv(output_file, sep="\t", index=True)
+    print(f"Saved filtered data to {output_file}")
+
+if __name__ == "__main__":
+    main()

### 2025-05-07 code/sustained_recovery.py

@@ -1,0 +1,12 @@
+import pandas as pd
+
+df = pd.read_csv('../results/filtered/anthro.tsv', sep='\t', index_col=0)
+
+t2 = df.query('(sampleID.str.contains("_52")) & (sampleID.str.contains("LCC1"))', engine='python')
+
+t2 = df.query('(sampleID.str.contains("_12")) & (sampleID.str.contains("LCC1"))', engine='python')
+
+t2 = df.query('(sampleID.str.contains("_0")) & (sampleID.str.contains("LCC1"))', engine='python')
+
+t2.WLZ_WHZ.gt(-1).sum()
+

### 2025-05-07 code/timetorecovery.py

@@ -1,0 +1,15 @@
+import pandas as pd
+import seaborn as sns
+from metatoolkit.change import change
+
+# Read the data
+df = pd.read_csv('../results/filtered/surveillance.tsv', sep='\t', index_col=0)
+meta = pd.read_csv('../results/filtered/meta.tsv', sep='\t', index_col=0)
+
+# Filter out rows where no days_to_recovery is available
+df.dropna(inplace=True)
+
+joined = df.join(meta)
+
+chan = change(df, meta.select_dtypes(include=[object]))
+joined.corr

### 2025-07-22 writing/manuscript/main.tex
Writing additions

@@
-% For styling
 \documentclass{article}
-\setlength\parindent{0pt}
-\setlength{\parskip}{\baselineskip}
-\usepackage[hidelinks]{hyperref}
+\usepackage{float}
+\usepackage[a4paper, top=2.5cm, bottom=2.5cm, left=2.5cm, right=2.5cm]{geometry}
+\usepackage[super,sort&compress]{natbib}
+\bibliographystyle{unsrtnat}
+\usepackage{textcomp}
+\usepackage{graphicx}
 \usepackage{authblk}
-\usepackage{graphicx}
-\usepackage{a4wide}
-\usepackage{textgreek}
 
-% For pandas dataframe
-\usepackage{booktabs}
-\usepackage{longtable}
+\title{
+  Understanding Recovery from Malnutrition in Early Life:\\
+  A Multi-Omic Analysis of the Infant Gut Microbiome, Plasma Lipidome, and Neurodevelopmental Outcomes
+}
 
-% For referencing
-%\let\cite\shortcite
-%\bibliographystyle{apacite}
-\bibliographystyle{plain}
-%\bibliographystyle{ksfh_nat}
-%%\usepackage[super]{natbib}
-%%\usepackage{hyperref}
-%%\bibliographystyle{unsrtnat}
+\author[1]{Authors}
+%\author[1,*]{Theo Portlock}
+%\author[2,*]{Talat Sharma}
+%\author[2,*]{Shahria Hafiz Kakon}
+%\author[3,*]{Berit Hartjen}
+%\author[1,*]{Chris Pook}
+%\author[1,*]{Brooke Wilson}
+%\author[3]{Ayisha Bhuttor}
+%\author[1]{Daniel Ho}
+%\author[1]{Inoli Shennon Wadumesthrige Don}
+%\author[3]{Anne-Michelle Engelstad}
+%\author[3]{Renata Di Lorenzo}
+%\author[3]{Garrett Greaves}
+%\author[3]{Caroline Kelsey}
+%\author[1]{Peter Gluckman}
+%\author[1,4,5,6]{Justin O'Sullivan}
+%\author[7]{Terrence Forrester}
+%\author[3]{Charles Nelson}
+%
+\affil[1]{Affiliations}
+%\affil[1]{The Liggins Institute, University of Auckland, NZ}
+%\affil[2]{Infectious Diseases Division, International Centre for Diarrheal Disease Research, Bangladesh}
+%\affil[3]{Department of Pediatrics, Boston Children’s Hospital and Harvard Medical School; Harvard Graduate School of Education, Boston, USA}
+%\affil[4]{The Maurice Wilkins Centre, The University of Auckland, New Zealand}
+%\affil[5]{MRC Lifecourse Epidemiology Unit, University of Southampton, UK}
+%\affil[6]{Singapore Institute for Clinical Sciences, A*STAR, Singapore}
+%\affil[7]{Faculty of Medical Sciences, UWI Solutions for Developing Countries, The University of the West Indies, Jamaica}
+%\affil[*]{These authors contributed equally}
 
-%\usepackage[backend=bibtex,
-%  style=nature,
-%  sorting=none,
-%  giveninits=true,
-%  isbn=false,doi=false,url=false,
-%  natbib=true
-%]{biblatex}
-%\addbibresource{library.bib}
-%\usepackage{hyperref}
-
-% For glossaries
-\usepackage[nonumberlist]{glossaries}
-\makeglossaries
-\newacronym{SHAP}{SHAP}{SHapley Additive exPlanations}
-\newacronym{MAM}{MAM}{Moderate Acute Malnutrition}
-\newacronym{EC}{EC}{Expressive Communication}
-\newacronym{EEG}{EEG}{Electroencephalography}
-\newacronym{AUCROC}{AUCROC}{Area Under the Curve}
-\newacronym{LMIC}{LMIC}{Low- and Middle- Income Countries}
-\newacronym{MWU}{MWU}{Mann-Whitney U test}
-\newacronym{WLZ/WHZ}{WLZ/WHZ}{weight-for-length/height}
-\newacronym{PSD}{PSD}{power spectral density}
-\newacronym{BBB}{BBB}{Blood Brain Barrier}
-\newacronym{MUAC}{MUAC}{Mid-upper arm circumference}
-\newacronym{PCoA}{PCoA}{Principal Coordinates Analysis}
-\newacronym{PERMANOVA}{PERMANOVA}{PERmutational Multivariate ANalysis Of VAriance}
-\newacronym{FDR}{FDR}{False Discovery Rate}
-\newacronym{HC}{HC}{Head Circumference}
-\newacronym{SCFA}{SCFA}{Short Chain Fatty Acid}
-\newacronym{P/B}{P/B}{\textit{Prevotella}-to-\textit{Bacteroides}}
-\newacronym{OCFA}{OCFA}{Odd Chain Fatty Acid}
-
-% For drafts
-\usepackage{setspace}
-\doublespacing
-\usepackage{lineno}
-\linenumbers
-
-% For captions at end of document
-%\usepackage{figcaps}
-
-\title{Linking the Gut Microbiome to Neurocognitive Development in Bangladesh Malnourished Infants}
-\author[1,*]{Theo Portlock}
-\author[2,*]{Talat Sharma}
-\author[2,*]{Shahria Hafiz Kakon}
-\author[3,*]{Berit Hartjen}
-\author[1,*]{Chris Pook}
-\author[1,*]{Brooke Wilson}
-\author[3]{Ayisha Bhuttor}
-\author[1]{Daniel Ho}
-\author[1]{Inoli Shennon Wadumesthrige Don}
-\author[3]{Anne-Michelle Engelstad}
-\author[3]{Renata Di Lorenzo}
-\author[3]{Garrett Greaves}
-\author[3]{Caroline Kelsey}
-\author[1]{Peter Gluckman}
-\author[1,4,5,6]{Justin O'Sullivan}
-\author[7]{Terrence Forrester}
-\author[3]{Charles Nelson}
-
-\affil[1]{The Liggins Institute, University of Auckland, NZ}
-\affil[2]{Infectious Diseases Division, International Centre for Diarrheal Disease Research, Bangladesh}
-\affil[3]{Department of Pediatrics, Boston Children’s Hospital and Harvard Medical School; Harvard Graduate School of Education, Boston, USA}
-\affil[4]{The Maurice Wilkins Centre, The University of Auckland, New Zealand}
-\affil[5]{MRC Lifecourse Epidemiology Unit, University of Southampton, University Road, Southampton, UK}
-\affil[6]{Singapore Institute for Clinical Sciences, Agency for Science Technology and Research, Singapore}
-\affil[7]{Faculty of Medical Sciences, UWI Solutions for Developing Countries, The University of the West Indies (UWI), Jamaica}
-\affil[*]{These authors contributed equally}
-\date{\vspace{-5ex}}
+\date{}
 
@@
 \maketitle
+
 \newpage
-\printglossaries
+\section*{Abstract}
+Recovery from undernutrition during childhood has historically centered on short term weight gain with little attention given to redressing the delays in neurocognitive development.
+Herein we discuss the results of a prospective, randomized controlled trial for the use of an enhanced, brain-targeted micronutrient supplemented nutritional rehabilitation regimen versus the standard of care and healthy age-matched controls on measures of cognition and emotional regulation in a cohort of one-year-old Bangladeshi children.
+Although the enhanced regimen had no impact on measures of beahviour (Bayley's, and Wolke's score), sleep, executive function (spin the pots, glitter wand), an investigation into the factors that influenced these cognitive factors revealed that genetic factors (Polygenetic Risk Score for Intelligence) were predictive (AUCROC 0.70) for expressive communication score at the one year baseline but lost it's predictive power at 24 months.
+Overall recovery was from MAM was more likely in infants with larger heads, higher baseline weight and higher weight change after 1 week.
+It was also more likely in those that had lower genetic potential for Non-Alcoholic Fatty Liver disease and higher Educational Attainment PRS scores as well as greater adherence to refeeding. 
+The enhanced regimen also demonstrated reduced days to recovery and an improvement in gut microbiome composition as exemplified by a depletion of \textit{Suterella wadsworthensis} and \textit{Lactobacillus fermentum}.
+Using integrated network analysis of mixed modelling and correlation between deltas reveal a potential connection that circulating proline concentration is correlated with EEG microstates (E Occurance) that influences EEG PSD measures in the brain as well as geneemotone and approach behavioural measurements.
+This longitudinal, multimodal study explains biological mechanism through network analysis to understand the impacts of a dietary intervention.
 
-\section*{Abstract}
-
+\newpage
 \section*{Main}
+Malnutrition is a global crisis (UN SDG2: Zero Hunger) \cite{UN_SDG2}.
+Malnutrition refers to deficiencies, excesses, or imbalances in a person’s intake of energy and/or nutrients and can refer to undernutrition, overweight, or obesity.
+Moderate Acute Malnutrition (MAM) is the most common, most treatable form of undernutrition.
+MAM in childhood delays the development of multiple systems in the body from the gut microbiome to neurocognition.
+A small proportion of children with MAM are treated.
+Treatment involves refeeding.
+Current Refeeding methods focus solely on short term weight gain, associate with serious adverse events (such as diarrhoea or Refeeding Syndrome), and are often ineffective at restoring developmental trajectories.
+Those that recover their weight often have lasting neurological deficits including cognition and emotional regulation.
+MAM is associated with gut microbiome immaturity.
+Novel prototypical refeeding methods, such as microbiota-directed complementary foods (MDCFs) target the gut-microbiome-brain axis as a mediator to address these deficits.
+By inoculating gnotobiotic pig and mice preclinical models with age-characteristic gut microorganisms, MDCFs increased the levels of biomarkers and mediators of growth, bone formation, neurodevelopment, and immune function toward a state resembling healthy children \cite{gehrig2019effects}.
+Developing these methods into sustainable refeeding strategies is needed urgently.
+Herein we discuss the results of a study that empoys The Enhanced Ready to Use Supplementary Food (E-RUTF) plus Small Quantity Lipid Based Nutrient Supplement (E-SQLNS) as compared to standard RUTF commonly used to treat childhood Moderate Acute Malnutrition (MAM)
 
@@
 As a city with the second highest density of population and in a country with childhood malnutrition rate is one of the highest globally, the Mirpur region in Dhaka, Bangladesh was chosen to assess the impact of early-life malnutrition \cite{ahmed2012nutrition}.
+Refeeding resulted in recovery in approximately half the children with MAM, with no significant difference in rate between the two refeed methods (43.0\% in the Local RUSF group vs 41.2\% in the ERUSF group , MWU.
+p = 0.9460).
+However, a key distinction emerged in time to recovery, which was significantly shorter by approximately ten days in the ERUSF group (p = 0.0422, d = 0.46) (Table 1)
 
-\begin{figure}[!htb]
+\begin{figure}[H]
 	\centering
-\includegraphics[scale=0.9]{../../figures/Figure1-microbiome.pdf}
-\caption[Malnutrition impacts the 12-month-old infant gut microbiome]{
-	Malnutrition impacts the 12-month-old infant gut microbiome.
-	a) Schematic of study design.
-	b) Summary of data collected.
-	c) Change in diversity of the gut microbiome associated with malnutrition.
-	d) PCoA Scatterplot of Bray-Curtis beta diversities of samples (each marker is a single infant sample).
-	e) Barplot of significant taxonomic differences in relative abundance and prevailance between 12-month-old well-nourished and \gls{MAM} samples (\textit{p} \textless{} 0.05).
-	f) Boxplot of \gls{P/B} ratio change between study conditions.
-	g) Volcano plot of pathways that associate with malnutrition (red and orange horizontal line signifies \textit{q} \textless{} 0.05 and 0.01 respectively.
-left and right vertical lines represent Log\textsubscript{2}(\gls{MAM}/well-nourished) of -0.1 and 0.1 respectively).}
+	\includegraphics[scale=0.3]{../../figures/table1.pdf}
+	\caption[Study population characteristics and recovery rates]{
+		Study population characteristics and recovery rates.
+	}
 \label{Figure1}
+\end{figure}
+
+\subsection*{Impact of the Enhanced Refeed regimen on Neurodevelopmental Outcomes}
+Using mixed-effects models controlling for sex and delivery mode, the enhanced refeed intervention showed no statistically significant main effects or interactions with timepoint across cognitive, communication, and motor Bayley scores (Table S2).
+Enhanced refeeding did not significantly improve most cognitive/behavioral outcomes, though some marginal trends in fine motor scores were observed.
+While children improved in their cognitive abilities over time, this improvement did not differ significantly by intervention group.
+
+\begin{figure}[H]
+\centering
+\includegraphics[scale=1.2]{../../figures/anthrochange_simple.pdf}
+	\caption[Measures of cognition were not significantly impacted by refeeding]{
+		Measures of cognition were not significantly impacted by refeeding.
+		a) Lineplot of how children change in WLZ/WHZ during refeeding.
+		b) Boxplot of days to recovery compared between the feeding groups.
+		c) Cognitive factors that were associated with refeeding (Left and right heatmaps of Bayley score and Wolkes respectively).
+		d) Lineplot of change of Expressive Communication Bayley Score stratified by Well-nourished, ERUSF, or RUSF.
+		e) Boxplot of change of vocalisation Wolkes score stratified by refeed type.
+	}
+\label{Figure2}
+\end{figure}
+
+\subsection*{Baseline factors that influenced Bayley scores}
+With the primary outcome investigated, a post-hoc, discovery approach into the factors that influenced cognition was performed.
+Random Forest regression analysis to predict Bayley scores at the baseline and two-year timepoint from the baseline datasets revealed that, although Bayley Scores were largely unexplainable from the baseline datasets, Polygenetic Risk Score (PRS) factors were predictive of expressive communication scores at the baseline timepoint.
+This predictive power was lost at the year-two timepoint.
+The PRS score that was most differentially associated with expressive communication score was the Intelligence PRS.
+\begin{figure}[H]
+	\centering
+	\includegraphics[scale=1.2]{../../figures/predictbayley.pdf}
+	\caption[Factors that were predictive of Bayley scores]{
+		Factors that were predictive of Bayley scores.
+	}
+\label{Figure1}
+\end{figure}
+
+\subsection*{Baseline Predictors of Recovery}
+Acknowleging that the refeed type made little impact on the overall recovery rates, an investigation into the factors that were associated with anthropometric recovery was undertaken.
+Anthropometric and genetic indicators were stronger predictors than microbiome or EEG.
+The most significant factors were anthropometric such as head circumference, indicating extent of malnutrition.
+Other factors included the proportion of days without feeding failure, indicating compliance/adherance to the refeeding.
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.8]{../../figures/ev.pdf}
+	\caption[Factors that associate with recovery.]{
+		Factors that associate with recovery.
+		Explained variance as measured by PERMANOVA of beta diversities between samples within each dataset.
+	}
+\label{Figure3}
+\end{figure}
+
+\subsection*{Assessing the impacts with respect to well-nourished controls}
+The classification of a given microbiome as 'healthy' has long been contested.
+This is true also of a number of datasets within this cohort.
+Thus, a method whereby the distance between the composition of the overall community of each dataset was compared with the respective well-nourished age-matched control for each timepoint.
+For each dataset analysed through this techhniqe, the shrinking/widening of this distance with respect to the well-nourished controls between the two timepoints was used to indicate improvement/deterioration.
+ERUTF drove a greater shift toward healthy microbiome, including depletion of S. wadsworthensis and L. fermentum.
+\begin{figure}[H]
+	\centering
+	\includegraphics[scale=0.6]{../../figures/newrecoveryfeed3.pdf}
+	\caption[Refeed methods improve/deteriorate bodily systems differently]{
+		Refeed methods improve/deteriorate bodily systems differently.
+		A, Radial plot of percentage change in distance from Well-nourished controls before and after refeeding (1yr and 2yr).
+Positive and negative percentages represent improvement and deterioration, respectively.
+		B, Horizontal barplot of features that explain the change significantly (q < 0.2) between refeed methods after mixed modelling for each dataset.
+		C, Heatmap to show clusters of dataset changes.
+		D, Boxplot to show how Lactobacillus fermentum changes according to the refeed method.
+	}
+\label{Figure1}
+\end{figure}
+
+\subsection*{Network-Based Integration of Multimodal Changes}
+With an understanding of the impacts that the enhanced refeed method had on developmental trajectories across the datasets recoreded, the next step was to understand how those changes might be connected together mechanisticly.
+It was found that proline metabolism was correlated with EEG and cognitive outcomes, suggesting neuro-metabolic recovery pathways.
+\begin{figure}[H]
+\centering
+\includegraphics[scale=0.5]{../../figures/mixedmodelnetwork.pdf}
+	\caption[Mechanism of E-RUTF differential impacts on brain structure]{
+		Mechanism of E-RUTF differential impacts on brain structure.
+	}
+\label{Figure4}
 \end{figure}
@@
 \section*{Methods}
+PSS method \cite{mozumder2022reliability}
 \subsection*{Ethics}
@@
 The census identified 5736 children aged between 11 to 13 months and 2,314 children aged between 34 to 38 months.
-During the census, if the guardian verbally consented to the study procedure, and the babies met the inclusion and exclusion criteria of the study (Table 1), the SWs proceeded to measure the \gls{MUAC} of the child.
-Mothers of babies who were within the \gls{MUAC} range were invited to visit the icddr,b study clinic for further assessment and enrolment.
+During the census, if the guardian verbally consented to the study procedure, and the babies met the inclusion and exclusion criteria of the study (Table 1), the SWs proceeded to measure the MUAC of the child.
+Mothers of babies who were within the MUAC range were invited to visit the icddr,b study clinic for further assessment and enrolment.
 Final screening for eligibility and study consent occurred at the icddr,b Mirpur study clinic.
@@
 Comprehension of the study was assessed using scripted points and open-ended questions.
-Following consent, the clinical screening team completed a screening form, capturing the date of enrolment, sex, date of birth (DOB), weight (in kg), length/ height (in cm), head circumference (in cm), and Mid-Upper Arm Circumference \gls{MUAC} measurements of the child.
-The \gls{WLZ/WHZ} Z-score for each child was calculated using the WHO anthropometric calculator.
+Following consent, the clinical screening team completed a screening form, capturing the date of enrolment, sex, date of birth (DOB), weight (in kg), length/ height (in cm), head circumference (in cm), and Mid-Upper Arm Circumference MUAC measurements of the child.
+The WLZ/WHZ Z-score for each child was calculated using the WHO anthropometric calculator.
 The child's age was validated using the EPI vaccination card.
@@
 Data was sampled at 500 Hz.
-Impedances were kept under 100 k \textomega{} when possible and measured once at the beginning of the session, and again halfway through.
+Impedances were kept under 100 k\textohm{} when possible and measured once at the beginning of the session, and again halfway through.
 Sessions were conducted in a dimly lit room with the participants sitting on the parent’s lap.
@@
 Relative abundances were adjusted by Centred Log Ratio to account for the compositional nature of the dataset \cite{gloor2016s}.
-Log adjusted fold change significance was measured using (\gls{MWU}) test using the ‘mannwhitneyu’ function from ‘scipy.stats’ and adjusted for multiple testing using the ‘fdrcorrection’ function from statsmodels.stats.multitest.
-\gls{PCoA} ordinations (plotted using 'skbio.stats.ordination.pcoa' module) were used to visualise the clustering of the Bray-Curtis dissimilarities (calculated using skbio.distance.pdist) between samples from their species and functional composition.
-To quantify the variance of the gut microbiome explained covariates, \gls{PERMANOVA} p-values were calculated from those Bray-Curtis Dissimilarities using the ’permanova’ function from the 'skbio.stats.distance' module.
+Log adjusted fold change significance was measured using (MWU) test using the ‘mannwhitneyu’ function from ‘scipy.stats’ and adjusted for multiple testing using the ‘fdrcorrection’ function from statsmodels.stats.multitest.
+PCoA ordinations (plotted using 'skbio.stats.ordination.pcoa' module) were used to visualise the clustering of the Bray-Curtis dissimilarities (calculated using skbio.distance.pdist) between samples from their species and functional composition.
+To quantify the variance of the gut microbiome explained covariates, PERMANOVA p-values were calculated from those Bray-Curtis Dissimilarities using the ’permanova’ function from the 'skbio.stats.distance' module.
 Bray-Curtis were also used to capture the temporal dynamics of the microbiome from baseline.
-Numerical Associations between species and metadata were measured with Spearman correlation (calculated using 'spearmanr' function from 'scipy.stats' module), where significance was defined as \gls{FDR} adjusted p-values of \textless{} 0.05 as per \citet{2020SciPyNMeth}.
+Numerical Associations between species and metadata were measured with Spearman correlation (calculated using 'spearmanr' function from 'scipy.stats' module), where significance was defined as FDR adjusted p-values of \textless{} 0.05 as per \citet{2020SciPyNMeth}.
 Associations between categorical data were measured with Fisher's Exact test (calculated using 'fisher\_exact' from 'scipy.stats' module), where significance was defined as p-values of \textless{} 0.05.
@@
 \subsection*{Network analysis}
-Absolute spearman rho of above 0.3 were used as edges and gut bacterial species and functional profiles, EEG, and plasma lipids were used as nodes coloured by their mean SHAP scores for classifier models that distinguish \gls{MAM} from well-nourished conditions.
+Absolute spearman rho of above 0.3 were used as edges and gut bacterial species and functional profiles, EEG, and plasma lipids were used as nodes coloured by their mean SHAP scores for classifier models that distinguish MAM from well-nourished conditions.
 
@@
 \bibliography{library}
-\printbibliography
+
+\newpage
 
 \section*{Supplementary material}
-
-\begin{table}[!htb]
+\begin{figure}[!htb]
 \centering
-\caption[Baseline infant characteristics]{
-	Baseline infant characteristics.
-	Plus minus values are means \textpm{} SD from continous variables and their pvalues are calculated using \gls{MWU}.
-	All other variables are categorical (True vs False) with their pvalues calculated using Fishers Exact test.}
-%\input{../../figures/texsupptables/TableS1_patientinfo}
-\label{TableS1}
-\end{table}
+\includegraphics[scale=0.8]{../../figures/newalldata.pdf}
+	\caption[Datasets recoreded in the M4EFaD study]{
+		Datasets recorded in the M4EFaD study.
+	}
+\label{Figure1}
+\end{figure}
 

### 2025-07-22 code/calculate_diversity.sh
Fix diversity

@@ -1,22 +1,33 @@
 #!/bin/bash
 # Biobakery's calculate diversity functions in their github/utils directory
+
+set -e
+
 export PATH="metaphlan/metaphlan/utils/:$PATH"
+export PATH="metatoolkit/metatoolkit/:$PATH"
 
 metaphlan_file=$1
+
 mkdir -p results/mbiome_alpha
+
+transpose.py ${metaphlan_file} -o results/mbiome_alpha/samplesascols.tsv
+
+faith_pd.py ${metaphlan_file} conf/mpa_vOct22_CHOCOPhlAnSGB_202212.nwk -o results/mbiome_alpha/faiths_pd.tsv
 
 # For alpha diversity measures
 for metric in gini shannon richness simpson
 do
-calculate_diversity.R \
-	-f ${metaphlan_file} \
-	-t conf/mpa_v31_CHOCOPhlAn_201901_species_tree.nwk \
+Rscript metaphlan/metaphlan/utils/calculate_diversity.R \
+	-f results/mbiome_alpha/samplesascols.tsv \
+	-t conf/mpa_vOct22_CHOCOPhlAnSGB_202212.nwk \
 	-d alpha \
 	-p alpha \
 	-m $metric \
-	-s s__ \
+	-s t__ \
 	-o results/mbiome_alpha
 done
+
+rm results/mbiome_alpha/samplesascols.tsv
 
 # Combine alpha diversity metrics and cleanup
 merge.py results/mbiome_alpha/* -o results/alpha_diversity.tsv &&

### 2025-07-22 code/extract_species.py

@@ -11,11 +11,8 @@ import pandas as pd
 # Load MetaPhlAn output with multi-index: (subjectID, timepoint)
 df = pd.read_csv('results/metaphlan.tsv', sep='\t', index_col=0)
 
-# Return to original shape: rows = samples, columns = taxa
-df = df.T
-
 # Keep only species-level taxa (those containing 's__')
-df = df.loc[:, df.columns.str.contains('s__')]
+df = df.loc[:, df.columns.str.contains('t__')]
 
 # Simplify species names: strip the full taxonomic path, keep only species name
 df.columns = df.columns.str.replace(r'.*\|s__', '', regex=True)
@@ -23,12 +20,6 @@ df.columns = df.columns.str.replace(r'.*\|s__', '', regex=True)
 # Label the columns as 'species'
 df.columns.name = 'species'
 
-# Transpose for normalization: rows = samples, columns = species
-df = df.T.div(df.sum(axis=1), axis=1)  # Normalize so each sample sums to 1
-
-# Transpose back to original shape: rows = samples, columns = species
-df = df.T
-
-# Export the normalized species-level table (flatten index for readability)
+# Save
 df.to_csv('results/species.tsv', sep='\t')
 

### 2025-07-22 code/format_metaphlan.py

@@ -3,13 +3,13 @@
 
 """
 Author: Theo Portlock
-Script to format MetaPhlAn3 output for downstream analysis
+Script to format MetaPhlAn4 output for downstream analysis
 """
 
 import pandas as pd
 
 # Load MetaPhlAn3 output file, skipping the first row (contains description/metadata)
-df = pd.read_csv('data/m4efad_metaphlan3_profiles_april2024.tsv',
+df = pd.read_csv('data/metaphlan_merged_profiles.tsv',
                  sep='\t', index_col=0, header=1)
 
 # Load the sample metadata
@@ -40,4 +40,4 @@ mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
 # Transmute and save formatted data - for alpha-diversity analysis
-df.T.to_csv('results/metaphlan.tsv', sep='\t')
+df.to_csv('results/metaphlan.tsv', sep='\t')

### 2025-07-22 writing/presentation/main.tex

\documentclass{beamer}
\usetheme{Madrid}
\usecolortheme{dolphin}
\usefonttheme{structurebold}

\title{Rethinking Refeeding: Cognitive Recovery from Childhood Undernutrition}
\author{Theo Portlock\inst{1} \and Justin O'Sullivan\inst{1}}
\institute{\inst{1}University of Auckland}
\date{03--04--2024}

\begin{document}

\frame{\titlepage}

\begin{frame}{Introduction}
\begin{itemize}
  \item Recovery
\end{itemize}
\end{frame}

\begin{frame}{SAP}
\begin{itemize}
  \item Plot recovery curves
  \item Use Fisher to look for recovery from baseline characteristics
  \item Use RF to look for recovery for each dataset
  \item Use SHAP to look for cross dataset interactions
\end{itemize}
\end{frame}

\begin{frame}{Anthropometric recovery from MAM was achieved in 44\% of children after refeeding}
\begin{center}
  \includegraphics[width=\linewidth]{../../figures/timerecovery.pdf}
\end{center}
\end{frame}

\begin{frame}{Recovery is associated with...}
\begin{center}
  \includegraphics[width=\linewidth]{../../figures/fisher.pdf}
\end{center}
\end{frame}

\begin{frame}{Recovery is difficult to predict}
\begin{itemize}
  \item One-hot encoding
\end{itemize}
\begin{center}
  \includegraphics[width=\linewidth]{figures/examplebrowser.pdf}
\end{center}
\end{frame}

\begin{frame}{An interaction predicts recovery}
\begin{center}
  \includegraphics[width=\linewidth]{figures/examplebrowser.pdf}
\end{center}
\end{frame}

\end{document}

### 2025-07-23 code/temp_predict.sh
mtk and manuscript

@@ -1,0 +1,69 @@
+
+#!/bin/bash
+# A tool for measuring the power of a model in capturing the importance of feature and feature interaction effects
+# Theo Portlock
+
+set +e
+
+source venv/bin/activate
+export PATH="code/:$PATH"
+export PATH="metatoolkit/metatoolkit/:$PATH"
+
+merge.py results/baseline/species.tsv \
+	results/baseline/aa.tsv
+	-o results/merged_aa_species.tsv
+
+SCIPY_ARRAY_API=1 test_train_split.py \
+	--input results/baseline/genetics.tsv \
+	--y_file results/filtered/metaonehot.tsv \
+	--y_col Recovery.Recovered \
+	--output_dir results/dataset_split
+
+random_forest.py \
+	--input_dir results/genetics_rf/ \
+	--task classification \
+	--output_model results/dataset_rf.pkl
+
+evaluate_model.py \
+	--model results/dataset_rf.pkl \
+	--input_dir results/dataset_split/ \
+	--task regression \
+	--report_file results/dataset_report_rf.tsv
+
+shap_interpret.py \
+	--model results/dataset_rf.pkl \
+	--input_dir results/dataset_split/ \
+	--shap_val \
+	--shap_interact \
+	--output_dir results/dataset_rf_shap 
+
+merge.py \
+	results/theoretical_variance/interaction_importance.tsv \
+	results/dataset_rf_shap/mean_abs_shap_interaction_test.tsv \
+	-a \
+	--add-filename \
+	-o results/merged_interactions.tsv
+
+format_merged.py
+
+plot.py
+
+create_network.py \
+	--edges results/format_merged_interactions.tsv \
+	--output results/network.graphml
+
+plot_network.py \
+	results/network.graphml \
+	--edge_color_attr mean_abs_shap_interaction_test.tsv \
+	--layout shell \
+	--cmap Reds \
+	--figsize 4 4 \
+	--output results/network2.svg
+
+shap_plots.sh
+
+arrange_svgs.py \
+	results/shap_plots_test_data/* \
+	--cols 2 \
+	--output results/shap_plots_merged.svg
+

### 2025-08-03 writing/manuscript/main.tex
mtk and manuscript
mtk and rda analysis

@@
 p = 0.9460).
-However, a key distinction emerged in time to recovery, which was significantly shorter by approximately ten days in the ERUSF group (p = 0.0422, d = 0.46) (Table 1)
+However, a key distinction emerged in time to recovery, which was shorter by approximately ten days in the ERUSF group (p = 0.0422, d = 0.46) (Table 1)
 
@@
 	\centering
-	\includegraphics[scale=0.3]{../../figures/table1.pdf}
+	\includegraphics[scale=0.3]{figures/table1.pdf}
 	\caption[Study population characteristics and recovery rates]{
@@
 Using mixed-effects models controlling for sex and delivery mode, the enhanced refeed intervention showed no statistically significant main effects or interactions with timepoint across cognitive, communication, and motor Bayley scores (Table S2).
-Enhanced refeeding did not significantly improve most cognitive/behavioral outcomes, though some marginal trends in fine motor scores were observed.
+Enhanced refeeding did not significantly improve most cognitive/b shavioral outcomes, though some marginal trends in fine motor scores were observed.
 While children improved in their cognitive abilities over time, this improvement did not differ significantly by intervention group.
@@
 \centering
-\includegraphics[scale=1.2]{../../figures/anthrochange_simple.pdf}
+\includegraphics[scale=1.2]{figures/anthrochange_simple.pdf}
 	\caption[Measures of cognition were not significantly impacted by refeeding]{
@@
 The PRS score that was most differentially associated with expressive communication score was the Intelligence PRS.
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=1.2]{../../figures/predictbayley.pdf}
-	\caption[Factors that were predictive of Bayley scores]{
-		Factors that were predictive of Bayley scores.
-	}
-\label{Figure1}
-\end{figure}
 
 \subsection*{Baseline Predictors of Recovery}
-Acknowleging that the refeed type made little impact on the overall recovery rates, an investigation into the factors that were associated with anthropometric recovery was undertaken.
+Acknowledging that the refeed type made little impact on the overall recovery rates, an investigation into the factors that were associated with anthropometric recovery was undertaken.
 Anthropometric and genetic indicators were stronger predictors than microbiome or EEG.
@@
 \centering
-\includegraphics[scale=0.8]{../../figures/ev.pdf}
+\includegraphics[scale=0.8]{figures/ev.pdf}
 	\caption[Factors that associate with recovery.]{
@@
 	\centering
-	\includegraphics[scale=0.6]{../../figures/newrecoveryfeed3.pdf}
+	\includegraphics[scale=0.6]{figures/newrecoveryfeed3.pdf}
 	\caption[Refeed methods improve/deteriorate bodily systems differently]{
@@
 \centering
-\includegraphics[scale=0.5]{../../figures/mixedmodelnetwork.pdf}
+\includegraphics[scale=0.5]{figures/mixedmodelnetwork.pdf}
 	\caption[Mechanism of E-RUTF differential impacts on brain structure]{
@@
 PSS method \cite{mozumder2022reliability}
+
 \subsection*{Ethics}
@@
 
-\bibliography{library}
+\bibliography{writing/manuscript/library}
 
@@
 \centering
-\includegraphics[scale=0.8]{../../figures/newalldata.pdf}
+\includegraphics[scale=0.8]{figures/newalldata.pdf}
 	\caption[Datasets recoreded in the M4EFaD study]{

### 2025-08-03 code/extract_genus.py

@@ -1,0 +1,26 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+
+"""
+Author: Theo Portlock
+Script to extract species-level relative abundances from MetaPhlAn output
+"""
+
+import pandas as pd
+
+# Load MetaPhlAn output with multi-index: (subjectID, timepoint)
+df = pd.read_csv('results/metaphlan.tsv', sep='\t', index_col=0)
+
+# Keep only species-level taxa (those containing 's__')
+df = df.loc[:, ~df.columns.str.contains('s__')]
+df = df.loc[:, df.columns.str.contains('g__')]
+
+# Simplify species names: strip the full taxonomic path, keep only species name
+df.columns = df.columns.str.replace(r'.*\|g__', '', regex=True)
+
+# Label the columns as 'genus'
+df.columns.name = 'genus'
+
+# Save
+df.to_csv('results/genus.tsv', sep='\t')
+

### 2025-08-03 test.sh

@@ -1,15 +1,0 @@
-maaslin3.R \
-  results/filtered/species.tsv \
-  results/timemeta.tsv \
-  results/maaslin/species_m3 \
-  --fixed_effects "Recovery,Condition" \
-  --reference "Recovery,No recovery;Condition,Well-nourished" \
-  --random_effects "subjectID" \
-  --min_prevalence 0.5 \
-  --small_random_effects TRUE \
-  --cores 10 \
-  --normalization TSS \
-  --transform LOG \
-  --standardize TRUE \
-  --plot_summary_plot FALSE \
-  --plot_associations FALSE

### 2025-08-03 code/anthrotime.py
Major changes

@@ -6,8 +6,8 @@ import seaborn as sns
 import pandas as pd
 
 # Load data
-anthro = pd.read_csv('results/anthro.tsv', sep='\t', index_col=0)
-timemeta = pd.read_csv('results/timemeta.tsv', sep='\t', index_col=0)
+anthro = pd.read_csv('results/filtered/anthro.tsv', sep='\t', index_col=0)
+timemeta = pd.read_csv('results/filtered/timemeta.tsv', sep='\t', index_col=0)
 df = anthro.join(timemeta)
 
 # Add condition to timemeta

### 2025-08-03 code/calculate_diversity.sh

@@ -7,6 +7,7 @@ export PATH="metaphlan/metaphlan/utils/:$PATH"
 export PATH="metatoolkit/metatoolkit/:$PATH"
 
 metaphlan_file=$1
+output_file=$2
 
 mkdir -p results/mbiome_alpha
 
@@ -30,5 +31,5 @@ done
 rm results/mbiome_alpha/samplesascols.tsv
 
 # Combine alpha diversity metrics and cleanup
-merge.py results/mbiome_alpha/* -o results/alpha_diversity.tsv &&
+merge.py results/mbiome_alpha/* -o $2 &&
 rm -r results/mbiome_alpha

### 2025-08-03 code/create_table1.py

@@ -7,6 +7,7 @@ from scipy.stats import chi2_contingency, f_oneway, ttest_ind
 from pathlib import Path
 from docx import Document
 from docx.shared import Inches
+import sys
 
 # === STATS FUNCTIONS ===
 def cohen_d(x, y):
@@ -180,7 +181,7 @@ for idx, row in summary_df.iterrows():
         row_cells[i + 1].text = str(val)
 
 # Save Word document
-word_output = Path("results/refeed_summary_with_stats.docx")
+word_output = Path(sys.argv[1])
 doc.save(word_output)
 print(f"Saved DOCX summary to {word_output}")
 

### 2025-08-03 code/extract_species.py

@@ -1,25 +1,0 @@
-#!/usr/bin/env python3
-# -*- coding: utf-8 -*-
-
-"""
-Author: Theo Portlock
-Script to extract species-level relative abundances from MetaPhlAn output
-"""
-
-import pandas as pd
-
-# Load MetaPhlAn output with multi-index: (subjectID, timepoint)
-df = pd.read_csv('results/metaphlan.tsv', sep='\t', index_col=0)
-
-# Keep only species-level taxa (those containing 's__')
-df = df.loc[:, df.columns.str.contains('t__')]
-
-# Simplify species names: strip the full taxonomic path, keep only species name
-df.columns = df.columns.str.replace(r'.*\|s__', '', regex=True)
-
-# Label the columns as 'species'
-df.columns.name = 'species'
-
-# Save
-df.to_csv('results/species.tsv', sep='\t')
-

### 2025-08-03 code/format_aa.py

@@ -63,6 +63,6 @@ mapping['sampleID'] = mapping['subjectID'] + '_' + mapping['timepoint'].astype(s
 mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
-df.to_csv('results/aa.tsv', sep='\t')
+df.to_csv('results/cleaned/aa.tsv', sep='\t')
 
 

### 2025-08-03 code/format_anthro.py

@@ -75,4 +75,4 @@ mapping['sampleID'] = mapping['subjectID'] + '_' + mapping['timepoint'].astype(s
 mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
-df.to_csv('results/anthro.tsv', sep='\t', index=True)
+df.to_csv('results/cleaned/anthro.tsv', sep='\t', index=True)

### 2025-08-03 code/format_bayley.py

@@ -35,4 +35,4 @@ mapping['sampleID'] = mapping['subjectID'] + '_' + mapping['timepoint'].astype(s
 mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
-df.to_csv('results/bayley.tsv', sep='\t')
+df.to_csv('results/cleaned/bayley.tsv', sep='\t')

### 2025-08-03 code/format_bfpob.py

@@ -14,4 +14,4 @@ df['PoB'] = df['H/O Place of birth (1=Home, 2 = Health Facility)'].map({1:'Home'
 df = df[['BF','PoB']]
 df.index.name = 'subjectID'
 
-df.to_csv('results/bfpob.tsv', sep='\t')
+df.to_csv('results/cleaned/bfpob.tsv', sep='\t')

### 2025-08-03 code/format_deliverysupp.py

@@ -15,4 +15,4 @@ df = df[['Delivery_Mode','Supplementation']]
 
 df.index.name = 'subjectID'
 
-df.to_csv('results/deliverysupp.tsv', sep='\t')
+df.to_csv('results/cleaned/deliverysupp.tsv', sep='\t')

### 2025-08-03 code/format_fcis.py

@@ -34,5 +34,5 @@ mapping['sampleID'] = mapping['subjectID'] + '_' + mapping['timepoint'].astype(s
 mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
-df.to_csv('results/fcis.tsv', sep='\t')
+df.to_csv('results/cleaned/fcis.tsv', sep='\t')
 

### 2025-08-03 code/format_feedsandrecovery.py

@@ -21,5 +21,5 @@ df.index.name='subjectID'
 df.loc[df.index.str[3] ==2, 'Feed'] = 'Healthy'
 df.loc[df.index.str[3] ==2, 'Recovery'] = 'Healthy'
 
-df.to_csv('results/feedandrecovery.tsv', sep='\t')
+df.to_csv('results/cleaned/feedandrecovery.tsv', sep='\t')
 

### 2025-08-03 code/format_fnirs.py

@@ -35,5 +35,5 @@ mapping['sampleID'] = mapping['subjectID'] + '_' + mapping['timepoint'].astype(s
 mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
-df.to_csv("results/fnirs.tsv", sep='\t')
+df.to_csv("results/cleaned/fnirs.tsv", sep='\t')
 

### 2025-08-03 code/format_genetics.py

@@ -27,4 +27,4 @@ joined = pd.concat(dfs, join='outer', axis=1).T.groupby(level=0).max().T
 joined.index.name = 'subjectID'
 
 # Save
-joined.to_csv('results/genetics.tsv', sep='\t')
+joined.to_csv('results/cleaned/genetics.tsv', sep='\t')

### 2025-08-03 code/format_glitter.py

@@ -34,5 +34,5 @@ mapping['sampleID'] = mapping['subjectID'] + '_' + mapping['timepoint'].astype(s
 mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
-df.to_csv('results/glitter.tsv', sep='\t')
+df.to_csv('results/cleaned/glitter.tsv', sep='\t')
 

### 2025-08-03 code/format_head.py

@@ -43,5 +43,5 @@ mapping['sampleID'] = mapping['subjectID'] + '_' + mapping['timepoint'].astype(s
 mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
-df.to_csv('results/head.tsv', sep='\t')
+df.to_csv('results/cleaned/head.tsv', sep='\t')
 

### 2025-08-03 code/format_humann.py

@@ -11,7 +11,7 @@ import pandas as pd
 
 # for functions - in cpm
 pathways = pd.read_csv("data/m4efad_humann3_pathway_cpm_april2024.tsv", index_col=0, sep='\t').T
-samplesheet = pd.read_csv('results/samplesheet.tsv', sep='\t', index_col=0)
+samplesheet = pd.read_csv('results/cleaned/samplesheet.tsv', sep='\t', index_col=0)
 
 pathways.index = pathways.index.str.replace('_Abundance','')
 #pathways = pathways.loc[:, ~pathways.columns.str.contains('UNMAPPED|UNINTEGRATED')]
@@ -26,7 +26,7 @@ df = df.loc[df.sum(axis=1) != 0, df.sum(axis=0) !=0]
 df.columns.name = 'stratified_pathway'
 #df = df.stack().to_frame('relative_abundance')
 #df = df.loc[df.relative_abundance != 0]
-df.to_csv('results/pathwaysstrat.tsv', sep='\t')
+df.to_csv('results/cleaned/pathwaysstrat.tsv', sep='\t')
 
 df = pathways.loc[:, ~pathways.columns.str.contains('\|')]
 df.columns = df.columns.str.replace(r'\:.*','', regex=True)
@@ -40,4 +40,4 @@ mapping['sampleID'] = mapping['subjectID'] + '_' + mapping['timepoint'].astype(s
 mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
-df.to_csv('results/pathways.tsv', sep='\t')
+df.to_csv('results/cleaned/pathways.tsv', sep='\t')

### 2025-08-03 code/format_kneaddata.py

@@ -9,7 +9,7 @@ import numpy as np
 import pandas as pd
 
 quality = pd.read_csv('data/m4efad_kneaddata_read_counts_april2024.tsv', sep='\t', index_col=0)
-samplesheet = pd.read_csv('results/samplesheet.tsv', sep='\t', index_col=0)
+samplesheet = pd.read_csv('results/cleaned/samplesheet.tsv', sep='\t', index_col=0)
 
 quality = quality.loc[:,quality.columns.str.contains('final')].sum(axis=1).div(1e6).to_frame('HQ read depth (million reads)')
 quality = quality.join(samplesheet.reset_index().set_index('Seq_ID')[['subjectID','timepoint']], how='inner').set_index(['subjectID','timepoint'])
@@ -20,4 +20,4 @@ mapping['sampleID'] = mapping['subjectID'] + '_' + mapping['timepoint'].astype(s
 mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
-df.to_csv('results/quality.tsv', sep='\t')
+df.to_csv('results/cleaned/quality.tsv', sep='\t')

### 2025-08-03 code/format_lipids.py

@@ -168,4 +168,4 @@ mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
 # Export the final data to a TSV file
-df.to_csv('results/lipids.tsv', sep='\t')
+df.to_csv('results/cleaned/lipids.tsv', sep='\t')

### 2025-08-03 code/format_metaphlan.py

@@ -13,7 +13,7 @@ df = pd.read_csv('data/metaphlan_merged_profiles.tsv',
                  sep='\t', index_col=0, header=1)
 
 # Load the sample metadata
-samplesheet = pd.read_csv('results/samplesheet.tsv',
+samplesheet = pd.read_csv('results/cleaned/samplesheet.tsv',
                           sep='\t', index_col=0)
 
 # Clean column names: remove ".metaphlan" suffix
@@ -40,4 +40,4 @@ mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
 # Transmute and save formatted data - for alpha-diversity analysis
-df.to_csv('results/metaphlan.tsv', sep='\t')
+df.to_csv('results/cleaned/metaphlan.tsv', sep='\t')

### 2025-08-03 code/format_micro.py

@@ -33,5 +33,5 @@ mapping['sampleID'] = mapping['subjectID'] + '_' + mapping['timepoint'].astype(s
 mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
-df.to_csv('results/micro.tsv', sep='\t')
+df.to_csv('results/cleaned/micro.tsv', sep='\t')
 

### 2025-08-03 code/format_microbiome_samplesheet.py

@@ -76,4 +76,4 @@ samplesheet.index = samplesheet.ID + samplesheet.timepoint
 # Rename
 samplesheet = samplesheet.rename(columns={'ID':'subjectID'})
 samplesheet.timepoint = samplesheet.timepoint.astype(int)
-samplesheet.to_csv('results/samplesheet.tsv', sep='\t', index=False)
+samplesheet.to_csv('results/cleaned/samplesheet.tsv', sep='\t', index=False)

### 2025-08-03 code/format_pci.py

@@ -38,4 +38,4 @@ mapping['sampleID'] = mapping['subjectID'] + '_' + mapping['timepoint'].astype(s
 mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
-df.to_csv('results/pci.tsv', sep='\t')
+df.to_csv('results/cleaned/pci.tsv', sep='\t')

### 2025-08-03 code/format_pots.py

@@ -29,5 +29,5 @@ df.columns = df.columns.str.replace(' ','_')
 # Drop timepoint as all 52
 df = df.droplevel(1)
 
-df.to_csv('results/pots.tsv', sep='\t')
+df.to_csv('results/cleaned/pots.tsv', sep='\t')
 

### 2025-08-03 code/format_psd.py

@@ -32,4 +32,4 @@ mapping['sampleID'] = mapping['subjectID'] + '_' + mapping['timepoint'].astype(s
 mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
-df.to_csv('results/psd.tsv', sep='\t')
+df.to_csv('results/cleaned/psd.tsv', sep='\t')

### 2025-08-03 code/format_pss.py

@@ -59,4 +59,4 @@ mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
 # Save final output
-df.to_csv('results/pss.tsv', sep='\t')
+df.to_csv('results/cleaned/pss.tsv', sep='\t')

### 2025-08-03 code/format_ses.py

@@ -52,7 +52,7 @@ sanitcols = df.columns[df.columns.str.contains('Washing')].tolist() + [
 ]
 sdf = df.loc[:, sanitcols]
 sdf.columns = sdf.columns.str.replace('Washing_-_', '')
-sdf.to_csv('results/sanitation.tsv', sep='\t')
+sdf.to_csv('results/cleaned/sanitation.tsv', sep='\t')
 df = df.loc[:, ~df.columns.isin(sanitcols)]
 
 # Define and extract household-related columns
@@ -68,7 +68,7 @@ hdf = df.loc[:, householdcols]
 hdf.columns = hdf.columns.str.replace('Household_-_','')
 hdf.columns = hdf.columns.str.replace('Table','Tabletop')
 hdf.columns = hdf.columns.str.replace('/', '_')
-hdf.to_csv('results/household.tsv', sep='\t')
+hdf.to_csv('results/cleaned/household.tsv', sep='\t')
 df = df.loc[:, ~df.columns.isin(householdcols)]
 
 # Define and extract family-related columns
@@ -80,7 +80,7 @@ familycols = [
     'Language'
 ]
 fdf = df.loc[:, familycols]
-fdf.to_csv('results/family.tsv', sep='\t')
+fdf.to_csv('results/cleaned/family.tsv', sep='\t')
 df = df.loc[:, ~df.columns.isin(familycols)]
 
 # Define and extract education-related columns
@@ -95,7 +95,7 @@ educationcols = [
     'Family_owns_the_home_they_live_in'
 ]
 edf = df.loc[:, educationcols]
-edf.to_csv('results/education.tsv', sep='\t')
+edf.to_csv('results/cleaned/education.tsv', sep='\t')
 df = df.loc[:, ~df.columns.isin(educationcols)]
 
 # Define and extract media consumption-related columns
@@ -106,7 +106,7 @@ mediacols = [
     'Uses_social_media'
 ]
 mdf = df.loc[:, mediacols]
-mdf.to_csv('results/media_consumption.tsv', sep='\t')
+mdf.to_csv('results/cleaned/media_consumption.tsv', sep='\t')
 df = df.loc[:, ~df.columns.isin(mediacols)]
 
 # Define and extract economic columns
@@ -116,8 +116,8 @@ ecdf.columns = ecdf.columns.str.replace('_(taka)','').str.replace('(','').str.re
 
[…]

### 2025-08-03 code/format_sleep.py

@@ -28,5 +28,5 @@ mapping['sampleID'] = mapping['subjectID'] + '_' + mapping['timepoint'].astype(s
 mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
-df.to_csv('results/sleep.tsv', sep='\t')
+df.to_csv('results/cleaned/sleep.tsv', sep='\t')
 

### 2025-08-03 code/format_surveillance.py

@@ -99,4 +99,4 @@ fail = b.rename(columns=mapping)
 surveillance = pd.concat([days_of_catchup, base_body_temp, avs, comp, fail], axis=1)
 
 # Save result
-surveillance.to_csv('results/surveillance.tsv', sep='\t')
+surveillance.to_csv('results/cleaned/surveillance.tsv', sep='\t')

### 2025-08-03 code/format_vep.py

@@ -36,4 +36,4 @@ mapping['sampleID'] = mapping['subjectID'] + '_' + mapping['timepoint'].astype(s
 mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
-df.to_csv('results/vep.tsv', sep='\t')
+df.to_csv('results/cleaned/vep.tsv', sep='\t')

### 2025-08-03 code/format_vitamin.py

@@ -40,4 +40,4 @@ mapping['sampleID'] = mapping['subjectID'] + '_' + mapping['timepoint'].astype(s
 mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
-df.to_csv('results/vitamin.tsv', sep='\t')
+df.to_csv('results/cleaned/vitamin.tsv', sep='\t')

### 2025-08-03 code/format_wolkes.py

@@ -36,4 +36,4 @@ mapping['sampleID'] = mapping['subjectID'] + '_' + mapping['timepoint'].astype(s
 mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
-df.to_csv("results/wolkes.tsv", sep='\t')
+df.to_csv("results/cleaned/wolkes.tsv", sep='\t')

### 2025-08-03 code/predict.py

@@ -1,170 +1,0 @@
-#!/usr/bin/env python3
-# -*- coding: utf-8 -*-
-
-from imblearn.over_sampling import SMOTE
-from itertools import permutations
-from pathlib import Path
-from sklearn.ensemble import RandomForestClassifier, RandomForestRegressor
-from sklearn.metrics import roc_auc_score, mean_absolute_error, r2_score, roc_curve
-from sklearn.model_selection import train_test_split
-from sklearn.preprocessing import StandardScaler
-import argparse
-import numpy as np
-import os
-import pandas as pd
-import shap
-import sys
-
-def load(subject):
-    if os.path.isfile(subject):
-        return pd.read_csv(subject, sep='\t', index_col=0)
-    return pd.read_csv(f'results/{subject}.tsv', sep='\t', index_col=0)
-
-def load_predictor(predictor_file, target_column):
-    if os.path.isfile(predictor_file):
-        df = pd.read_csv(predictor_file, sep='\t', index_col=0)
-    elif os.path.isfile(f'results/{predictor_file}.tsv'):
-        df = pd.read_csv(f'results/{predictor_file}.tsv', sep='\t', index_col=0)
-    else:
-        raise FileNotFoundError(f"Predictor file {predictor_file} not found.")
-
-    if target_column not in df.columns:
-        raise ValueError(f"Target column '{target_column}' not found in {predictor_file}.")
-
-    return df[target_column]
-
-def save(df, subject, index=True):
-    output_path = f'results/{subject}.tsv'
-    os.makedirs(os.path.dirname(output_path), exist_ok=True)
-    df.to_csv(output_path, sep='\t', index=index)
-
-def predict(df, y, analysis, subject, shap_val=False, shap_interact=False, n_iter=30):
-    outputs = []
-    aucrocs = []
-    fpr_tpr = []
-    maes = []
-    r2s = []
-    meanabsshaps = pd.DataFrame()
-    shap_interacts = pd.DataFrame(index=pd.MultiIndex.from_tuples(permutations(df.columns, 2)))
-
-    random_state = 1
-    for random_state in range(n_iter):
-        if analysis.lower() == 'classifier':
-            model = RandomForestClassifier(
-                n_jobs=-1, random_state=random_state,
-                bootstrap=False, max_depth=None,
-                min_samples_leaf=8, min_samples_split=20, n_estimators=400
-            )
-        elif analysis.lower() == 'regressor':
-            model = RandomForestRegressor(
-                n_jobs=-1, random_state=random_state
[…]

### 2025-08-03 code/timetorecovery.py

@@ -1,18 +1,0 @@
-#!/usr/bin/env python3
-# -*- coding: utf-8 -*-
-
-import pandas as pd
-import seaborn as sns
-from metatoolkit.change import change
-
-# Read the data
-df = pd.read_csv('results/filtered/surveillance.tsv', sep='\t', index_col=0)
-meta = pd.read_csv('results/filtered/meta.tsv', sep='\t', index_col=0)
-
-# Filter out rows where no days_to_recovery is available
-df.dropna(inplace=True)
-
-joined = df.join(meta)
-
-chan = change(df, meta.select_dtypes(include=[object]))
-joined.corr

### 2025-08-05 code/format_lipids.py
Lots of changes

@@ -168,4 +168,4 @@ mapping = mapping[['sampleID', 'subjectID', 'timepoint']]
 df.index = mapping['sampleID']
 
 # Export the final data to a TSV file
-df.to_csv('results/cleaned/lipids.tsv', sep='\t')
+df.to_csv('results/cleaned/unlogged_lipids.tsv', sep='\t')

### 2025-08-05 code/mofa_export.py

@@ -1,0 +1,67 @@
+#!/usr/bin/env python
+
+import mofax as mofa
+import muon as mu
+import pandas as pd
+import os
+import argparse
+
+# --- Parse arguments ---
+parser = argparse.ArgumentParser(description="Export MOFA+ factors and loadings")
+parser.add_argument('--model', default='results/mofa/t1.hdf5')
+parser.add_argument('--mdata', default='results/mofa/mdata.h5mu')
+parser.add_argument('--outdir', default='results/mofa')
+args = parser.parse_args()
+
+os.makedirs(args.outdir, exist_ok=True)
+
+# --- Load model and mdata ---
+model = mofa.mofa_model(args.model)
+mdata = mu.read(args.mdata)
+
+# --- Export factors ---
+factors_arr = model.get_factors()
+sample_index = model.get_samples()
+
+# Check shape match
+if len(sample_index) != factors_arr.shape[0]:
+    print(f"Error: Sample count mismatch. {len(sample_index)} sample names vs {factors_arr.shape[0]} factor rows.")
+    exit()
+
+# Handle tuple-style sample names like ('group', 'LCC1001')
+if isinstance(sample_index[0], tuple):
+    sample_index = [i[1] for i in sample_index]
+
+factors_df = pd.DataFrame(
+    factors_arr,
+    index=sample_index.iloc[:, 1],
+    columns=[f"Factor_{i+1}" for i in range(model.nfactors)]
+)
+factors_df.to_csv(os.path.join(args.outdir, 'sample_factors.tsv'), sep='\t')
+print("Exported sample factors.")
+
+# --- Export loadings ---
+weights = model.get_weights()
+if not isinstance(weights, dict):
+    weights = {model.get_views()[0]: weights}
+
+for view_name, weights_arr in weights.items():
+    try:
+        features = mdata[view_name].var_names
+    except KeyError:
+        print(f"Skipping {view_name}, features not found.")
+        continue
+
+    if weights_arr.shape[0] != len(features):
+        print(f"Skipping {view_name}, shape mismatch.")
+        print(f"  weights_arr.shape[0] = {weights_arr.shape[0]}, len(features) = {len(features)}")
+        continue
+
+    df = pd.DataFrame(
+        weights_arr,
+        index=features,
+        columns=[f"Factor_{i+1}" for i in range(model.nfactors)]
+    )
+    df.to_csv(os.path.join(args.outdir, f'loadings_{view_name}.tsv'), sep='\t')
+    print(f"Exported loadings for {view_name}")
+

### 2025-08-05 code/mofa_load_data.py

@@ -1,0 +1,35 @@
+#!/usr/bin/env python
+
+import pandas as pd
+import scanpy as sc
+import muon as mu
+from pathlib import Path
+import os
+import argparse
+from glob import glob
+
+parser = argparse.ArgumentParser(description="Load modalities and metadata into MuData object")
+parser.add_argument('--data-dir', default='results/timepoints/yr1')
+parser.add_argument('--meta-file', default='results/filtered/meta.tsv')
+parser.add_argument('--output', default='results/mofa/mdata.h5mu')
+args = parser.parse_args()
+
+Path(os.path.dirname(args.output)).mkdir(parents=True, exist_ok=True)
+
+print("Loading metadata...")
+obs = pd.read_csv(args.meta_file, sep='\t', index_col=0)
+
+print("Loading modalities...")
+data_files = glob(f'{args.data_dir}/*')
+if not data_files:
+    print(f"Error: No data files found in {args.data_dir}")
+    exit()
+
+mods = {Path(f).stem: sc.AnnData(pd.read_csv(f, sep='\t', index_col=0)) for f in data_files}
+mdata = mu.MuData(mods)
+mdata.obs = mdata.obs.join(obs, how='inner')
+
+print("Saving MuData object...")
+mdata.write(args.output)
+print(f"Saved MuData to {args.output}")
+

### 2025-08-05 code/mofa_plot_r2.py

@@ -1,0 +1,33 @@
+#!/usr/bin/env python
+
+import mofax as mofa
+import argparse
+import matplotlib.pyplot as plt
+import seaborn as sns
+import pandas as pd
+import numpy as np
+
+parser = argparse.ArgumentParser(description="Visualize R² from MOFA+ model")
+parser.add_argument('--model', default='results/mofa/t1.hdf5')
+args = parser.parse_args()
+
+print("Loading MOFA model...")
+model = mofa.mofa_model(args.model)
+r2_long_df = model.get_variance_explained()
+
+# Barplot
+plt.figure(figsize=(10, 6))
+sns.barplot(data=r2_long_df, x='View', y='R2', estimator=np.sum, errorbar=None)
+plt.title('Total R² per View')
+plt.xticks(rotation=45, ha='right')
+plt.tight_layout()
+plt.show()
+
+# Heatmap
+r2_df = r2_long_df.pivot_table(index='View', columns='Factor', values='R2')
+plt.figure(figsize=(12, 10))
+sns.heatmap(r2_df, cmap='viridis', vmax=15, annot=True, fmt=".2f")
+plt.title('R² per View and Factor')
+plt.tight_layout()
+plt.show()
+

### 2025-08-05 code/mofa_run.py

@@ -1,0 +1,25 @@
+#!/usr/bin/env python
+
+import muon as mu
+import os
+import argparse
+
+parser = argparse.ArgumentParser(description="Run MOFA+ on MuData object")
+parser.add_argument('--input', default='results/mofa/mdata.h5mu')
+parser.add_argument('--output', default='results/mofa/t1.hdf5')
+parser.add_argument('--factors', type=int, default=15)
+args = parser.parse_args()
+
+os.makedirs(os.path.dirname(args.output), exist_ok=True)
+
+print("Running MOFA+...")
+mdata = mu.read(args.input)
+mu.tl.mofa(
+    mdata,
+    use_obs='union',
+    n_factors=args.factors,
+    convergence_mode='medium',
+    outfile=args.output
+)
+print(f"MOFA+ model saved to {args.output}")
+

### 2025-08-05 code/mofanew.py

@@ -1,0 +1,135 @@
+#!/usr/bin/env python
+
+import pandas as pd
+import scanpy as sc
+import muon as mu
+import mofax as mofa
+import os
+from glob import glob
+from pathlib import Path
+import matplotlib.pyplot as plt
+import seaborn as sns
+import numpy as np
+
+# --- 1. Setup and Data Loading ---
+DATA_DIR = 'results/timepoints/yr1'
+META_FILE = 'results/filtered/meta.tsv'
+OUTPUT_DIR = 'results/mofa'
+MOFA_MODEL_FILE = os.path.join(OUTPUT_DIR, 't1.hdf5')
+
+Path(OUTPUT_DIR).mkdir(parents=True, exist_ok=True)
+
+try:
+    obs = pd.read_csv(META_FILE, sep='\t', index_col=0)
+except FileNotFoundError:
+    print(f"Error: Metadata file not found at {META_FILE}")
+    exit()
+
+print("Loading data modalities...")
+data_files = glob(f'{DATA_DIR}/*')
+if not data_files:
+    print(f"Error: No data files found in {DATA_DIR}")
+    exit()
+
+mods = {Path(i).stem: sc.AnnData(pd.read_csv(i, sep='\t', index_col=0)) for i in data_files}
+mdata = mu.MuData(mods)
+print(f"MuData object created with views: {list(mdata.mod.keys())}")
+mdata.obs = mdata.obs.join(obs, how='inner')
+print("Metadata joined to MuData object.")
+
+# --- 2. MOFA+ Analysis ---
+print("\nRunning MOFA+ analysis...")
+mu.tl.mofa(
+    mdata,
+    use_obs='union',
+    n_factors=15,
+    convergence_mode='medium',
+    outfile=MOFA_MODEL_FILE
+)
+print(f"MOFA+ analysis complete. Model saved to {MOFA_MODEL_FILE}")
+
+try:
+    model = mofa.mofa_model(MOFA_MODEL_FILE)
+except FileNotFoundError:
+    print(f"Error: MOFA model not found at {MOFA_MODEL_FILE}")
+    exit()
+
+# --- 3. Visualization and Analysis ---
+print("\nPerforming visualizations...")
+
+# Use the model object to get the R^2 values as a long-format DataFrame
+r2_long_df = model.get_variance_explained()
+
+# Plot R^2 per view using seaborn, which is more reliable
+print("Plotting R² per view...")
[…]

### 2025-08-05 code/mofatmp.py

@@ -1,0 +1,70 @@
+#!/usr/bin/env python
+
+import pandas as pd
+import io
+import scanpy as sc
+import muon as mu
+import numpy as np
+import pandas as pd
+import mofax as mofa
+import os
+from glob import glob
+from pathlib import Path
+import matplotlib.pyplot as plt
+
+data_dir = 'results/timepoints/yr1'
+obs = pd.read_csv('results/filtered/meta.tsv', sep='\t', index_col=0)
+
+data = glob(f'{data_dir}/*')
+mods = {Path(i).stem:sc.AnnData(pd.read_csv(i, sep='\t', index_col=0)) for i in data}
+
+mdata = mu.MuData(mods)
+mdata.obs = mdata.obs.join(obs)
+
+mu.tl.mofa(mdata,
+           use_obs='union',
+           n_factors=15,
+           convergence_mode='medium',
+           outfile='results/mofa/t1.hdf5')
+
+model = mofa.mofa_model("results/mofa/t1.hdf5")
+
+mofa.plot_r2(model, x='View', vmax=15)
+
+mu.pl.mofa_r2(mdata, x='View', vmax=15)  # ← this is what you want
+
+mu.pl.mofa(mdata, color='Recovery')
+#mu.pp.intersect_obs(mdata)
+
+print(model.factors.keys())     # should list factors
+print(model.weights.keys())     # should list views
+
+print(model.r2_per_view)
+
+import pandas as pd
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+# Extract view names
+views = list(mdata.uns['mofa']['variance'].keys())
+
+# Extract R² values per view and assemble into a DataFrame
+r2_dict = {
+    view: mdata.uns['mofa']['variance'][view]
+    for view in views
+}
+
+# Create DataFrame with views as rows and factors as columns
+r2_df = pd.DataFrame(r2_dict).T  # transpose so rows = views
+
+# Optional: label the factor columns
+r2_df.columns = [f"Factor {i}" for i in range(r2_df.shape[1])]
+
+# Plot the heatmap
+sns.heatmap(r2_df, cmap='viridis', vmax=15)
+plt.title('R² per View and Factor')
+plt.xlabel('Factors')
+plt.ylabel('Views')
+plt.tight_layout()
+plt.show()
+

### 2025-08-08 code/mofa_export.py
mtk and mofa

@@ -28,15 +28,16 @@ if len(sample_index) != factors_arr.shape[0]:
     print(f"Error: Sample count mismatch. {len(sample_index)} sample names vs {factors_arr.shape[0]} factor rows.")
     exit()
 
-# Handle tuple-style sample names like ('group', 'LCC1001')
-if isinstance(sample_index[0], tuple):
-    sample_index = [i[1] for i in sample_index]
+# Extract just the 'sample' column if sample_index is a DataFrame
+if isinstance(sample_index, pd.DataFrame):
+    sample_index = sample_index.iloc[:, 1]  # or sample_index["sample"]
 
 factors_df = pd.DataFrame(
-    factors_arr,
-    index=sample_index.iloc[:, 1],
-    columns=[f"Factor_{i+1}" for i in range(model.nfactors)]
-)
+       factors_arr,
+       index=sample_index,
+       columns=[f"Factor_{i+1}" for i in range(model.nfactors)]
+       )
+
 factors_df.to_csv(os.path.join(args.outdir, 'sample_factors.tsv'), sep='\t')
 print("Exported sample factors.")
 

### 2025-08-08 code/mofa_plots.py

@@ -16,7 +16,7 @@ model = mofa.mofa_model(args.model)
 r2_long_df = model.get_variance_explained()
 
 # Barplot
-plt.figure(figsize=(10, 6))
+plt.figure(figsize=(4, 3))
 sns.barplot(data=r2_long_df, x='View', y='R2', estimator=np.sum, errorbar=None)
 plt.title('Total R² per View')
 plt.xticks(rotation=45, ha='right')
@@ -25,8 +25,8 @@ plt.show()
 
 # Heatmap
 r2_df = r2_long_df.pivot_table(index='View', columns='Factor', values='R2')
-plt.figure(figsize=(12, 10))
-sns.heatmap(r2_df, cmap='viridis', vmax=15, annot=True, fmt=".2f")
+plt.figure(figsize=(4, 3))
+sns.heatmap(r2_df, cmap='viridis', vmax=15)
 plt.title('R² per View and Factor')
 plt.tight_layout()
 plt.show()

### 2025-08-08 code/temp_predict.sh

@@ -9,25 +9,21 @@ source venv/bin/activate
 export PATH="code/:$PATH"
 export PATH="metatoolkit/metatoolkit/:$PATH"
 
-merge.py results/baseline/species.tsv \
-	results/baseline/aa.tsv
-	-o results/merged_aa_species.tsv
-
 SCIPY_ARRAY_API=1 test_train_split.py \
-	--input results/baseline/genetics.tsv \
-	--y_file results/filtered/metaonehot.tsv \
-	--y_col Recovery.Recovered \
+	--input results/mofa/sample_factors.tsv \
+	--y_file results/filtered/meta.tsv \
+	--y_col Recovery \
 	--output_dir results/dataset_split
 
 random_forest.py \
-	--input_dir results/genetics_rf/ \
+	--input_dir results/dataset_split/ \
 	--task classification \
 	--output_model results/dataset_rf.pkl
 
 evaluate_model.py \
 	--model results/dataset_rf.pkl \
 	--input_dir results/dataset_split/ \
-	--task regression \
+	--task classification \
 	--report_file results/dataset_report_rf.tsv
 
 shap_interpret.py \
@@ -36,29 +32,6 @@ shap_interpret.py \
 	--shap_val \
 	--shap_interact \
 	--output_dir results/dataset_rf_shap 
-
-merge.py \
-	results/theoretical_variance/interaction_importance.tsv \
-	results/dataset_rf_shap/mean_abs_shap_interaction_test.tsv \
-	-a \
-	--add-filename \
-	-o results/merged_interactions.tsv
-
-format_merged.py
-
-plot.py
-
-create_network.py \
-	--edges results/format_merged_interactions.tsv \
-	--output results/network.graphml
-
-plot_network.py \
-	results/network.graphml \
-	--edge_color_attr mean_abs_shap_interaction_test.tsv \
-	--layout shell \
-	--cmap Reds \
-	--figsize 4 4 \
-	--output results/network2.svg
 
 shap_plots.sh
 

### 2025-08-10 writing/manuscript/main.tex
manuscript updates

@@
 \usepackage{authblk}
+\usepackage{booktabs} % Recommended for cleaner, professional-looking tables
 
@@
 
+\tableofcontents
+
 \newpage
-\section*{Abstract}
+\section{Abstract}
 Recovery from undernutrition during childhood has historically centered on short term weight gain with little attention given to redressing the delays in neurocognitive development.
 Herein we discuss the results of a prospective, randomized controlled trial for the use of an enhanced, brain-targeted micronutrient supplemented nutritional rehabilitation regimen versus the standard of care and healthy age-matched controls on measures of cognition and emotional regulation in a cohort of one-year-old Bangladeshi children.
-Although the enhanced regimen had no impact on measures of beahviour (Bayley's, and Wolke's score), sleep, executive function (spin the pots, glitter wand), an investigation into the factors that influenced these cognitive factors revealed that genetic factors (Polygenetic Risk Score for Intelligence) were predictive (AUCROC 0.70) for expressive communication score at the one year baseline but lost it's predictive power at 24 months.
-Overall recovery was from MAM was more likely in infants with larger heads, higher baseline weight and higher weight change after 1 week.
-It was also more likely in those that had lower genetic potential for Non-Alcoholic Fatty Liver disease and higher Educational Attainment PRS scores as well as greater adherence to refeeding. 
-The enhanced regimen also demonstrated reduced days to recovery and an improvement in gut microbiome composition as exemplified by a depletion of \textit{Suterella wadsworthensis} and \textit{Lactobacillus fermentum}.
-Using integrated network analysis of mixed modelling and correlation between deltas reveal a potential connection that circulating proline concentration is correlated with EEG microstates (E Occurance) that influences EEG PSD measures in the brain as well as geneemotone and approach behavioural measurements.
 This longitudinal, multimodal study explains biological mechanism through network analysis to understand the impacts of a dietary intervention.
@@
 \newpage
-\section*{Main}
+\section{Main}
+% Importance of MAM
 Malnutrition is a global crisis (UN SDG2: Zero Hunger) \cite{UN_SDG2}.
+% Definition of MAM
 Malnutrition refers to deficiencies, excesses, or imbalances in a person’s intake of energy and/or nutrients and can refer to undernutrition, overweight, or obesity.
+% Why MAM and not SAM
 Moderate Acute Malnutrition (MAM) is the most common, most treatable form of undernutrition.
+% What does MAM do
 MAM in childhood delays the development of multiple systems in the body from the gut microbiome to neurocognition.
+% Treament rates
 A small proportion of children with MAM are treated.
+% What is treatment
 Treatment involves refeeding.
+% Problems with refeeding
 Current Refeeding methods focus solely on short term weight gain, associate with serious adverse events (such as diarrhoea or Refeeding Syndrome), and are often ineffective at restoring developmental trajectories.
 Those that recover their weight often have lasting neurological deficits including cognition and emotional regulation.
+% Microbiome connection
 MAM is associated with gut microbiome immaturity.
+% Microbiome targetted refeeding
 Novel prototypical refeeding methods, such as microbiota-directed complementary foods (MDCFs) target the gut-microbiome-brain axis as a mediator to address these deficits.
+% Effectiveness of MCDF
 By inoculating gnotobiotic pig and mice preclinical models with age-characteristic gut microorganisms, MDCFs increased the levels of biomarkers and mediators of growth, bone formation, neurodevelopment, and immune function toward a state resembling healthy children \cite{gehrig2019effects}.
+% Need to improve on these methods
 Developing these methods into sustainable refeeding strategies is needed urgently.
+% The difficulty in predicting Recovery measurements
+% Our study aims
 Herein we discuss the results of a study that empoys The Enhanced Ready to Use Supplementary Food (E-RUTF) plus Small Quantity Lipid Based Nutrient Supplement (E-SQLNS) as compared to standard RUTF commonly used to treat childhood Moderate Acute Malnutrition (MAM)
 
-\section*{Results}
-\subsection*{Study population characteristics}
+\section{Results}
+\subsection{Study population characteristics}
+% Why was Bangladesh picked
 As a city with the second highest density of population and in a country with childhood malnutrition rate is one of the highest globally, the Mirpur region in Dhaka, Bangladesh was chosen to assess the impact of early-life malnutrition \cite{ahmed2012nutrition}.
+% Number of children selected and when (marasmic vs kwashiokor wasn't defined)
+% Recovery statistics (rates, days to recovery, and sustained recovery with biases measured) - \ref{Table1}. Complications (feeding failure, days of disease)
 Refeeding resulted in recovery in approximately half the children with MAM, with no significant difference in rate between the two refeed methods (43.0\% in the Local RUSF group vs 41.2\% in the ERUSF group , MWU.
@@
 
-\begin{figure}[H]
+\begin{table}[h!]
 	\centering
-	\includegraphics[scale=0.3]{figures/table1.pdf}
-	\caption[Study population characteristics and recovery rates]{
-		Study population characteristics and recovery rates.
-	}
-\label{Figure1}
-\end{figure}
-
-\subsection*{Impact of the Enhanced Refeed regimen on Neurodevelopmental Outcomes}
-Using mixed-effects models controlling for sex and delivery mode, the enhanced refeed intervention showed no statistically significant main effects or interactions with timepoint across cognitive, communication, and motor Bayley scores (Table S2).
-Enhanced refeeding did not significantly improve most cognitive/b shavioral outcomes, though some marginal trends in fine motor scores were observed.
-While children improved in their cognitive abilities over time, this improvement did not differ significantly by intervention group.
+	\caption{Study population characteristics, sociodemographic information, comorbitities, and recovery rates}
+	%\input{my_imported_table.tex}
+	\label{Table1}
+\end{table}
 
@@
 		b) Boxplot of days to recovery compared between the feeding groups.
-		c) Cognitive factors that were associated with refeeding (Left and right heatmaps of Bayley score and Wolkes respectively).
-		d) Lineplot of change of Expressive Communication Bayley Score stratified by Well-nourished, ERUSF, or RUSF.
-		e) Boxplot of change of vocalisation Wolkes score stratified by refeed type.
 	}
-\label{Figure2}
+	\label{Figure1}
 \end{figure}
 
-\subsection*{Baseline factors that influenced Bayley scores}
-With the primary outcome investigated, a post-hoc, discovery approach into the factors that influenced cognition was performed.
-Random Forest regression analysis to predict Bayley scores at the baseline and two-year timepoint from the baseline datasets revealed that, although Bayley Scores were largely unexplainable from the baseline datasets, Polygenetic Risk Score (PRS) factors were predictive of expressive communication scores at the baseline timepoint.
-This predictive power was lost at the year-two timepoint.
-The PRS score that was most differentially associated with expressive communication score was the Intelligence PRS.
+\subsection{What was associated with recovery}
+% A background to the data collected
+Acknowledging that the refeed type made little impact on the overall recovery rates, an investigation into the factors that were associated with anthropometric recovery was undertaken.
+% risk factors for sustained recovery
+% are the same factors important for reocvery rates/days to recovery/sustained recovery?
+% Why mixed models - maaslin3
 
-\subsection*{Baseline Predictors of Recovery}
-Acknowledging that the refeed type made little impact on the overall recovery rates, an investigation into the factors that were associated with anthropometric recovery was undertaken.
-Anthropometric and genetic indicators were stronger predictors than microbiome or EEG.
+% Influence of anthropometrics on recovery
 The most significant factors were anthropometric such as head circumference, indicating extent of malnutrition.
-Other factors included the proportion of days without feeding failure, indicating compliance/adherance to the refeeding.
-\begin{figure}[H]
-\centering
-\includegraphics[scale=0.8]{figures/ev.pdf}
-	\caption[Factors that associate with recovery.]{
-		Factors that associate with recovery.
-		Explained variance as measured by PERMANOVA of beta diversities between samples within each dataset.
-	}
-\label{Figure3}
-\end{figure}
+% Influence of genetics on recovery
+%% Background on the PRS method
+% Influence of behaviour on recovery
+%% Introduce measurements (FCIS, Glitter wand, Spin the pots, PCI, PSS, Sleep, 
+% Influence of socioeconomic environment on recovery
+Other factors included the proportion of days without feeding failure, and comorbidites
+% Influence of faecal microbiome on recovery
+%% alpha diversity
+%% beta diversity
+%% species
+%% functions
+% Influence of plasma metabolites on recovery
+%% AA
+%% Vitamins
 
-\subsection*{Assessing the impacts with respect to well-nourished controls}
+\subsection{Factors improved/deteriorated during the treatment}
+% Definition of improvement
 The classification of a given microbiome as 'healthy' has long been contested.
 This is true also of a number of datasets within this cohort.
+% Explaination of approach taken from the RDA version!
 Thus, a method whereby the distance between the composition of the overall community of each dataset was compared with the respective well-nourished age-matched control for each timepoint.
-For each dataset analysed through this techhniqe, the shrinking/widening of this distance with respect to the well-nourished controls between the two timepoints was used to indicate improvement/deterioration.
-ERUTF drove a greater shift toward healthy microbiome, including depletion of S. wadsworthensis and L. fermentum.
+For each dataset analysed through this technique, the shrinking/widening of this distance with respect to the well-nourished controls between the two timepoints was used to indicate improvement/deterioration.
+% approach results
+ERUTF drove a greater shift toward healthy microbiome, including depletion of \textit{S. wadsworthensis} and \textit{L. fermentum}.
+
 \begin{figure}[H]
@@
 	}
-\label{Figure1}
+\label{Figure2}
 \end{figure}
 
-\subsection*{Network-Based Integration of Multimodal Changes}
+\subsection{Using a systems biology approach to understand recovery}
+% Difficulties of interpretation of multiple datasets - including missing data
+% MOFA and factorisation
+% Machine learning pipeline explaination
+
+\subsection{Network-Based Integration of Multimodal Changes to understand recovery mechanism}
+% Networks as a tool to understand data relationships
+% Limitations of causality
+% How to link the results together
 With an understanding of the impacts that the enhanced refeed method had on developmental trajectories across the datasets recoreded, the next step was to understand how those changes might be connected together mechanisticly.
+% Network stats - number of nodes, edges, directional nodes, interaction effects
 It was found that proline metabolism was correlated with EEG and cognitive outcomes, suggesting neuro-metabolic recovery pathways.
+
 \begin{figure}[H]
@@
 	}
-\label{Figure4}
+\label{Figure3}
 \end{figure}
 
-\section*{Discussion}
+\section{Discussion}
 
-\section*{Conclusion}
+\section{Conclusion}
 
-\section*{Methods}
+\section{Methods}
 PSS method \cite{mozumder2022reliability}
 
-\subsection*{Ethics}
+\subsection{Ethics}
 The M4EFaD intervention was registered NCT05629624 on clinicaltrials.gov.
@@
 
-\subsection*{Study Design and Participants}
-The study was performed on the baseline data from three cohorts of infants who were enrolled (between Jan – December 2022) as part of the M4EFaD intervention within the Mirpur slum, Dhaka, Bangladesh.
+\subsection{Study Design and Participants}
+The study was performed on the baseline, and two year data from three cohorts of infants who were enrolled (between Jan – December 2022) as part of the M4EFaD intervention within the Mirpur slum, Dhaka, Bangladesh.
 Inclusion criteria included a diagnosis of malnutrition, no history of chronic medical conditions, and no antibiotic use within the past month.
 
-\subsection*{Recruitment and anthropometric data collection}
-Enrolment was initiated on February 7, 2022, and will continue until February 2024.
+\subsection{Recruitment and anthropometric data collection}
+Enrolment was initiated on February 7, 2022, and continued until February 2024.
 Study surveillance workers (SWs) conducted a door-to-door census (approximately 100,000 households) in Mirpur DNCC wards ward 2, 3 and 5 between January and December 2022.
@@
 The census identified 5736 children aged between 11 to 13 months and 2,314 children aged between 34 to 38 months.
-During the census, if the guardian verbally consented to the study procedure, and the babies met the inclusion and exclusion criteria of the study (Table 1), the SWs proceeded to measure the MUAC of the child.
+During the census, if the guardian verbally consented to the study procedure, and the babies met the inclusion and exclusion criteria of the study, the SWs proceeded to measure the MUAC of the child.
 Mothers of babies who were within the MUAC range were invited to visit the icddr,b study clinic for further assessment and enrolment.
@@
 
-\subsection*{EEG data collection and analysis}
+\subsection{EEG data collection and analysis}
 Continuous scalp EEG was recorded using NetStation 4.5.4. and 128-channel Hydrocel Geodesic Sensor Nets modified to remove eye electrodes (Electrical Geodesics, Inc. (EGI), Eugene, OR, USA).
@@
 
-\subsection*{Developmental Outcomes} 
+\subsection{Developmental Outcomes} 
 The Bayley Scales of Infant and Toddler Development, Fourth Edition (BSID-IV) cognitive, language, and motor subscales were administered to all participants.
@@
 
-\subsection*{Biological sample collection}
+\subsection{Biological sample collection}
 Stool samples were collected from each infant at their home at the baseline visit.
@@
 
-\subsection*{Microbiome DNA extraction and sequencing}
+\subsection{Microbiome DNA extraction and sequencing}
 DNA was extracted from stool samples using the ZymoBIOMICS MagBead DNA/RNA extraction kit (Zymo Research, \#R2136) following the standard protocol.
@@
 
-\subsection*{Plasma lipidomics}
+\subsection{Plasma lipidomics}
 Plasma samples for lipidomics were thawed on ice and extracted according to a method modified from \citet{liu2016plasma}.
@@
 
-\subsection*{Statistical Analyses}
+\subsection{Statistical Analyses}
 Python version 3.9.2 was used to perform all analysis \cite{van1995python}.
@@
 
-\subsection*{Machine learning}
+\subsection{Machine learning}
 SHAP Value (SHapley Additive exPlanations) interpretation was used to interpret the contributions each feature had on the model's performance using the ‘shap’ python package \cite{lundberg2017unified}.
 
-\subsection*{Network analysis}
+\subsection{Network analysis}
 Absolute spearman rho of above 0.3 were used as edges and gut bacterial species and functional profiles, EEG, and plasma lipids were used as nodes coloured by their mean SHAP scores for classifier models that distinguish MAM from well-nourished conditions.
 
-\section*{Code availability}
+\section{Code availability}
 All analysis code is available on the GitHub repository.
@@
 
-\section*{Ethics approval and consent to participate}
+\section{Ethics approval and consent to participate}
 Ethical approvals were obtained from the Research Review Committee (RRC; August 21, 2021) and Ethical Review Committee (ERC) of icddr,b (protocol no: PR-21084; September 21, 2021), Institutional Review Board of Boston Children’s Hospital, USA (for analyses of neuropsychological assessments), University of Auckland, New Zealand (approval AH23922; for analyses of collected biological samples) and University of West Indies (CREC-MN.51, 21/22).
 
-\section*{Data availability}
+\section{Data availability}
 EEG and metadata are available from the authors, upon reasonable request that meets the ethics of the study.
 
-\section*{Competing interests}
+\section{Competing interests}
 The authors declare that they have no competing interests.
 
-\section*{Funding}
+\section{Funding}
 Work on this clinical trial is supported by Wellcome Leap (9942 Culver Blvd Unit 1277 Culver City, CA 90232-4167, United States; www.wellcomeleap.org) to PDG, JMO, TF and CAN as part of the 1kD Program.
@@
 
-\section*{Author Contributions}
+\section{Author Contributions}
 TP, KG and JOS drafted and co-wrote the manuscript.
@@
 
-\section*{Acknowledgements}
+\section{Acknowledgements}
 The authors would like to acknowledge the participants in Mirpur, Dhaka, Bangladesh for their contributions to this study.
@@
 
-\section*{Supplementary material}
+\section{Supplementary material}
 \begin{figure}[!htb]

### 2025-08-11 code/sustained_recovery.py
Manuscript updates

@@ -3,12 +3,13 @@
 
 import pandas as pd
 
-df = pd.read_csv('results/filtered/anthro.tsv', sep='\t', index_col=0)
+df = pd.read_csv('results/timepoints/yr2/anthro.tsv', sep='\t', index_col=0)
 meta = pd.read_csv('results/filtered/meta.tsv', sep='\t', index_col=0)
 
-t2 = df.query('(sampleID.str.contains("_52")) & (sampleID.str.contains("LCC1"))', engine='python')
+# Just recovered
+recovered = meta.loc[meta.Recovery == 'Recovered']
+sus_recovered = df.loc[recovered.index, 'WLZ_WHZ'].gt(-1)
+meta['Sus_recovered'] = sus_recovered
 
-t2 = t2.loc[t2.WLZ_WHZ.gt(-1)]
 
-t2.index  = t2.index.str.replace('_52', '', regex=False)
-t2['sus_recovery'] = True
+

### 2025-08-13 code/mofa_load_data.py
Updating healthy schematic and prompt

@@ -9,7 +9,12 @@ import argparse
 from glob import glob
 
 parser = argparse.ArgumentParser(description="Load modalities and metadata into MuData object")
-parser.add_argument('--data-dir', default='results/timepoints/yr1')
+parser.add_argument('--t1-dir', default='results/timepoints/yr1',
+                    help='Directory containing T1 (baseline) data files')
+#parser.add_argument('--t2-dir', default='results/timepoints/yr2',
+#                    help='Directory containing T2 (after refeeding) data files')
+parser.add_argument('--notp-dir', default='results/timepoints/notp',
+                    help='Directory containing data with no timepoint info (optional)')
 parser.add_argument('--meta-file', default='results/filtered/meta.tsv')
 parser.add_argument('--output', default='results/mofa/mdata.h5mu')
 args = parser.parse_args()
@@ -19,14 +24,31 @@ Path(os.path.dirname(args.output)).mkdir(parents=True, exist_ok=True)
 print("Loading metadata...")
 obs = pd.read_csv(args.meta_file, sep='\t', index_col=0)
 
-print("Loading modalities...")
-data_files = glob(f'{args.data_dir}/*')
-if not data_files:
-    print(f"Error: No data files found in {args.data_dir}")
-    exit()
+print("Loading modalities for all timepoints...")
+mods = {}
 
-mods = {Path(f).stem: sc.AnnData(pd.read_csv(f, sep='\t', index_col=0)) for f in data_files}
+def load_timepoint_data(tp_name, tp_dir):
+    """Helper to load all data files from a given directory into AnnData objects."""
+    data_files = glob(f'{tp_dir}/*')
+    if not data_files:
+        print(f"Warning: No data files found in {tp_dir}")
+        return {}
+    tp_mods = {}
+    for f in data_files:
+        view_name = f"{Path(f).stem}_{tp_name}"  # keep timepoint in view name if needed
+        adata = sc.AnnData(pd.read_csv(f, sep='\t', index_col=0))
+        adata.obs['timepoint'] = tp_name
+        tp_mods[view_name] = adata
+    return tp_mods
+
+mods.update(load_timepoint_data('t1', args.t1_dir))
+#mods.update(load_timepoint_data('t2', args.t2_dir))
+mods.update(load_timepoint_data('notp', args.notp_dir))
+
+# Merge into MuData
 mdata = mu.MuData(mods)
+
+# Join metadata
 mdata.obs = mdata.obs.join(obs, how='inner')
 
 print("Saving MuData object...")

### 2025-08-14 writing/manuscript/main.tex
Manuscript updates
Updated figure 1

@@
 \usepackage{authblk}
-\usepackage{booktabs} % Recommended for cleaner, professional-looking tables
+\usepackage{booktabs}
 
@@
 \section{Main}
+% Definition of MAM
+Malnutrition refers to deficiencies, excesses, or imbalances in a person’s intake of energy and/or nutrients and can refer to undernutrition, overweight, or obesity.
 % Importance of MAM
 Malnutrition is a global crisis (UN SDG2: Zero Hunger) \cite{UN_SDG2}.
-% Definition of MAM
-Malnutrition refers to deficiencies, excesses, or imbalances in a person’s intake of energy and/or nutrients and can refer to undernutrition, overweight, or obesity.
 % Why MAM and not SAM
@@
 
-\subsection{What was associated with recovery}
+\subsection{What influenced recovery}
 % A background to the data collected
@@
 
+\subsection{What did recovery influence}
+% Maaslin3
+
 \subsection{Factors improved/deteriorated during the treatment}
@@
 Positive and negative percentages represent improvement and deterioration, respectively.
-		B, Horizontal barplot of features that explain the change significantly (q < 0.2) between refeed methods after mixed modelling for each dataset.
+		B, Horizontal barplot of features that explain the change significantly (q \textless{} 0.2) between refeed methods after mixed modelling for each dataset.
 		C, Heatmap to show clusters of dataset changes.

### 2025-08-14 code/anthrotime.py

@@ -47,7 +47,7 @@ def line(df, x, y, hue, ax):
 fig, axs = plt.subplots(1, 3, figsize=(6, 2.5), sharex=True)
 
 # Plot each metric
-line(df, 'timepoint', 'WLZ_WHZ', 'Recovery', axs[0])
+line(df, 'timepoint', 'WLZ_WHZ', 'Feed', axs[0])
 line(df, 'timepoint', 'MUAC', 'Recovery', axs[1])
 line(df, 'timepoint', 'Weight', 'Recovery', axs[2])
 

### 2025-08-14 code/format_surveillance.py

@@ -50,7 +50,7 @@ df = df.set_index(['subjectID', 'timepoint'])
 
 # Work out days to recovery
 days_of_catchup = df.reset_index().groupby(['Food','subjectID'])['Day'].last().xs(1)
-#days_of_catchup[days_of_catchup == 90.0] = np.nan
+days_of_catchup[days_of_catchup == 90.0] = np.nan
 days_of_catchup = days_of_catchup.rename('days_of_catchup')
 
 ## quick and dirty analyis

### 2025-08-14 code/schematic.py

@@ -1,0 +1,82 @@
+#!/usr/bin/env python
+# Create an improved study timeline subplot as an SVG/PNG
+import matplotlib.pyplot as plt
+from matplotlib.patches import Rectangle, FancyArrowPatch
+from matplotlib.lines import Line2D
+
+# Figure
+fig, ax = plt.subplots(figsize=(10, 5), dpi=200)
+
+ax.set_xlim(0, 10)
+ax.set_ylim(-1, 6)
+ax.axis('off')
+
+# Helper to draw a row
+def draw_row(y, label_left, label_right, t1_label="12 mo (T1)", t2_label="24 mo (T2)", n=None, intervention_label=None):
+    # timeline
+    ax.add_line(Line2D([1, 9], [y, y]))
+    # sample points
+    ax.plot([2, 8], [y, y], marker='s', linestyle='None', markersize=10)
+    # labels for timepoints
+    ax.text(2, y+0.35, t1_label, ha='center', va='bottom', fontsize=9)
+    ax.text(8, y+0.35, t2_label, ha='center', va='bottom', fontsize=9)
+    # left/right group labels
+    ax.text(0.2, y, label_left, ha='left', va='center', fontsize=10)
+    if label_right:
+        ax.text(9.3, y, label_right, ha='left', va='center', fontsize=10)
+    # n
+    if n is not None:
+        ax.text(0.2, y-0.45, f"n={n}", ha='left', va='center', fontsize=9)
+    # intervention band
+    if intervention_label:
+        rect = Rectangle((2.25, y-0.25), 5.5, 0.5, alpha=0.15, lw=0)
+        ax.add_patch(rect)
+        ax.text(5.0, y, intervention_label, ha='center', va='center', fontsize=9)
+        # direction arrow (optional visual flow)
+        arrow = FancyArrowPatch((2.25, y-0.35), (7.75, y-0.35), arrowstyle='-|>', mutation_scale=8, lw=1)
+        ax.add_patch(arrow)
+
+# Rows
+draw_row(4.5, "MAM cohort", "ERUSF + ESQLNS", n=70, intervention_label="Enhanced refeed & supplement")
+draw_row(3.0, "MAM cohort", "RUSF + SQLNS", n=70, intervention_label="Local refeed & supplement")
+draw_row(1.5, "Well-nourished controls", "", n=70)
+
[…]

### 2025-08-18 writing/manuscript/main.tex
Progress for the mefisto run

@@
 
-\newpage
 \section{Abstract}
@@
 
-\newpage
 \section{Main}
@@
 	\centering
-	\includegraphics[scale=1.2]{figures/anthrochange_simple.pdf}
-	\caption[Measures of cognition were not significantly impacted by refeeding]{
-		Measures of cognition were not significantly impacted by refeeding.
-		a) Lineplot of how children change in WLZ/WHZ during refeeding.
-		b) Boxplot of days to recovery compared between the feeding groups.
+	\includegraphics[scale=1.2]{figures/recovery2.pdf}
+	\caption[Study design and changes in weight-for-length z-scores (WLZ) in children with moderate acute malnutrition (MAM) receiving enhanced or local refeeding]{
+		Study design and changes in weight-for-length z-scores (WLZ) in children with moderate acute malnutrition (MAM) receiving enhanced or local refeeding.
+		a) study design for the evaluation of children with MAM between 12 and 24 months of age, randomized to either an enhanced refeed and supplement arm (ERUSF followed by ESQLNS) or a local refeed and supplement arm (RUSF followed by SQLNS), alongside a well-nourished control group.
+Baseline (T1) and post-refeeding (T2) sampling included anthropometry, stool microbiome, and plasma metabolites, with additional one time measures of household, educational, and genetic covariates.
+Recovery was defined as WLZ \textless{} –1 SD.
+		b) individual trajectories of WLZ over 52 weeks, with rapid catch-up growth during the first 15 weeks and stratification by recovery status.
+		c) proportion of participants achieving recovery, sustaining recovery at 24 months, or showing no recovery in each MAM treatment arm.
+		d) distribution of time to recovery between treatment arms, with a significant difference (\textit{q} \textless{} 0.05) between local RUSF and ERUSF groups.
+Box plots show the median, the 25th - 75th percentiles (IQR), and whiskers extending to 1.5 x IQR.
 	}
@@
 Acknowledging that the refeed type made little impact on the overall recovery rates, an investigation into the factors that were associated with anthropometric recovery was undertaken.
+% Prediction of recovery from baseline facors - MOFA
 % risk factors for sustained recovery

### 2025-08-18 code/mefisto_load.py

@@ -1,0 +1,53 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+
+import argparse
+import pandas as pd
+from pathlib import Path
+
+def main():
+    parser = argparse.ArgumentParser(description="Merge multiple omics with timemeta")
+    parser.add_argument("--timemeta", required=True, help="Path to timemeta table (TSV)")
+    parser.add_argument("--omics", nargs="+", required=True, help="List of omics TSV files")
+    parser.add_argument("--out", required=True, help="Output merged TSV (long format)")
+    args = parser.parse_args()
+
+    # Read timemeta (index = sampleID, keep subjectID and metadata)
+    timemeta = pd.read_csv(args.timemeta, sep="\t", index_col=0)
+
+    merged_long = []
+
+    for f in args.omics:
+        fpath = Path(f)
+        view = fpath.stem  # use filename (e.g. "head" from head.tsv)
+        df = pd.read_csv(f, sep="\t", index_col=0)  # sampleID as index
+
+        # Join with timemeta (align by sampleID)
+        df = timemeta.join(df, how="inner")
+
+        # Melt into long format
+        df_long = df.reset_index().melt(
+            id_vars=["index", "subjectID", "timepoint"] + [c for c in timemeta.columns if c not in ["subjectID", "timepoint"]],
+            var_name="feature",
+            value_name="value"
+        )
+
+        df_long["view"] = view
+        merged_long.append(df_long)
+
+    # Concatenate all omics into one long table
+    final = pd.concat(merged_long, ignore_index=True)
+
+    # Rename columns
+    final = final.rename(columns={"subjectID": "group"})
+    final = final.rename(columns={"index": "sample"})
+    print(final)
+
+    # Save
+    final.to_csv(args.out, sep="\t", index=False)
+
+if __name__ == "__main__":
+    main()
+
+
+from mofapy2.run.entry_point import entry_point

### 2025-08-18 code/mefisto_run.py

@@ -1,0 +1,34 @@
+#!/usr/bin/env python
+from mofapy2.run.entry_point import entry_point
+import pandas as pd
+
+# Load long-format input
+df = pd.read_csv("results/mofa/mefisto_input.tsv", sep='\t')
+
+ent = entry_point()
+ent.set_data_options(center_groups=False)
+ent.set_data_df(df)
+
+# --- Create samples_metadata for each group ---
+groups = df['group'].unique()
+samples_metadata = []
+for g in groups:
+    # select unique rows for this group, with sampleID and covariates
+    tmp = df[df['group']==g][['sample', 'timepoint']].drop_duplicates().set_index('sample')
+    samples_metadata.append(tmp)
+
+ent.data_opts["samples_metadata"] = samples_metadata
+
+# Now set covariates
+ent.set_covariates("timepoint", covariates_names="timepoint")
+
+# Model options
+ent.set_model_options(factors=10)
+ent.set_train_options(seed=42)
+ent.set_smooth_options(n_grid=10, start_opt=50, opt_freq=50)
+
+# Build, run, save
+ent.build()
+ent.run()
+ent.save("results/mofa/mofa_model.hdf5")
+

### 2025-08-18 code/mofa_run.py

@@ -16,9 +16,10 @@ print("Running MOFA+...")
 mdata = mu.read(args.input)
 mu.tl.mofa(
     mdata,
-    use_obs='union',
+    use_obs='union', # Fills in the missing pieces
     n_factors=args.factors,
     convergence_mode='medium',
+    gpu_mode=True,
     outfile=args.output
 )
 print(f"MOFA+ model saved to {args.output}")

### 2025-08-18 code/sustained_recovery.py

@@ -2,6 +2,8 @@
 # -*- coding: utf-8 -*-
 
 import pandas as pd
+import seaborn as sns
+import matplotlib.pyplot as plt
 
 df = pd.read_csv('results/timepoints/yr2/anthro.tsv', sep='\t', index_col=0)
 meta = pd.read_csv('results/filtered/meta.tsv', sep='\t', index_col=0)
@@ -11,5 +13,71 @@ recovered = meta.loc[meta.Recovery == 'Recovered']
 sus_recovered = df.loc[recovered.index, 'WLZ_WHZ'].gt(-1)
 meta['Sus_recovered'] = sus_recovered
 
+counts = meta.reset_index().groupby(['Feed','Recovery','Sus_recovered']).count()['subjectID'].reset_index()
 
+# meta must already have: Feed, Recovery, Sus_recovered
+def classify_recovery(row):
+    if row['Recovery'] == 'Recovered' and row['Sus_recovered']:
+        return 'Sustained recovery'
+    elif row['Recovery'] == 'Recovered':
+        return 'Recovered'
+    else:
+        return 'No recovery'
+
+meta['Recovery_status'] = meta.apply(classify_recovery, axis=1)
+
+# --- Count and percentage calculation ---
+status_order = ['Recovered', 'Sustained recovery', 'No recovery']
+colors = {
+    'Recovered': '#91bfdb',           # light blue
+    'Sustained recovery': '#0571b0',  # darker blue
+    'No recovery': '#d73027'          # red
+}
+
+# Count per Feed × Recovery_status
+counts = (
+    meta.groupby(['Feed', 'Recovery_status'])
+    .size()
+    .reset_index(name='Count')
+    .pivot(index='Feed', columns='Recovery_status', values='Count')
+    .fillna(0)
+    .reindex(columns=status_order)  # ensure order
+)
+
+# Percentages
+percentages = counts.div(counts.sum(axis=1), axis=0) * 100
+
+# --- Plotting ---
+fig, ax = plt.subplots(figsize=(2.6, 1.5))
+
+bottom = None
+for status in status_order:
+    ax.bar(
+        counts.index,
+        percentages[status],
+        bottom=bottom,
+        color=colors[status],
+        label=status
+    )
+
+    # Annotate counts in the middle of each section
+    for i, (pct, cnt) in enumerate(zip(percentages[status], counts[status])):
+        if pct > 0:
+            y_pos = (bottom[i] if bottom is not None else 0) + pct / 2
[…]

### 2025-08-18 code/mefisto_load.py
Prev parallel attempt for join for mefisto

@@ -4,17 +4,20 @@
 import argparse
 import pandas as pd
 from pathlib import Path
+from mofapy2.run.entry_point import entry_point
 
-def main():
+
+def parse_args():
     parser = argparse.ArgumentParser(description="Merge multiple omics with timemeta")
     parser.add_argument("--timemeta", required=True, help="Path to timemeta table (TSV)")
     parser.add_argument("--omics", nargs="+", required=True, help="List of omics TSV files")
     parser.add_argument("--out", required=True, help="Output merged TSV (long format)")
-    args = parser.parse_args()
+    return parser.parse_args()
 
+
+def main(args):
     # Read timemeta (index = sampleID, keep subjectID and metadata)
     timemeta = pd.read_csv(args.timemeta, sep="\t", index_col=0)
-
     merged_long = []
 
     for f in args.omics:
@@ -27,7 +30,8 @@ def main():
 
         # Melt into long format
         df_long = df.reset_index().melt(
-            id_vars=["index", "subjectID", "timepoint"] + [c for c in timemeta.columns if c not in ["subjectID", "timepoint"]],
+            id_vars=["index", "subjectID", "timepoint"]
+            + [c for c in timemeta.columns if c not in ["subjectID", "timepoint"]],
             var_name="feature",
             value_name="value"
         )
@@ -39,15 +43,14 @@ def main():
     final = pd.concat(merged_long, ignore_index=True)
 
     # Rename columns
-    final = final.rename(columns={"subjectID": "group"})
-    final = final.rename(columns={"index": "sample"})
+    final = final.rename(columns={"subjectID": "group", "index": "sample"})
     print(final)
 
     # Save
     final.to_csv(args.out, sep="\t", index=False)
 
+
 if __name__ == "__main__":
-    main()
+    args = parse_args()
+    main(args)
 
-
-from mofapy2.run.entry_point import entry_point

### 2025-08-18 code/mefrun.py

@@ -1,0 +1,78 @@
+#!/usr/bin/env python
+
+import argparse
+import pandas as pd
+from mofapy2.run.entry_point import entry_point
+
+
+def parse_args():
+    parser = argparse.ArgumentParser(
+        description="Run MEFISTO on multi-omic input data with optional covariates"
+    )
+    parser.add_argument(
+        "-i", "--input",
+        required=True,
+        help="Input TSV file containing merged omics data (with SampleID, SubjectID, Timepoint)"
+    )
+    parser.add_argument(
+        "-o", "--output",
+        default="results/mofa/mofa_model.hdf5",
+        help="Output path for the trained MEFISTO model (HDF5 format)"
+    )
+    parser.add_argument(
+        "--factors",
+        type=int,
+        default=10,
+        help="Number of factors to train (default: 10)"
+    )
+    parser.add_argument(
+        "--seed",
+        type=int,
+        default=42,
+        help="Random seed for reproducibility (default: 42)"
+    )
+    parser.add_argument(
+        "--n_grid",
+        type=int,
+        default=10,
+        help="Number of grid points for temporal smoothing (default: 10)"
+    )
+    return parser.parse_args()
+
+
+def main(args):
+    # Load merged data
+    df = pd.read_csv(args.input, sep="\t")
+
+    # Setup MEFISTO entry point
+    ent = entry_point()
+    ent.set_data_options(
+        center_groups=False,
+        scale_groups=False,
+        likelihoods="gaussian"
+    )
+    ent.set_data_df(df, sample_id_name="SampleID", group_id_name="SubjectID")
+
+    # Add covariates if present
+    if "timepoint" in df.columns:
+        ent.set_covariates(sample_covariates=df[["SampleID", "timepoint"]],
+                           covariates_names="timepoint")
+
+    # Model and training options
+    ent.set_model_options(factors=args.factors)
+    ent.set_train_options(seed=args.seed)
+    ent.set_smooth_options(n_grid=args.n_grid, start_opt=50, opt_freq=50)
+
+    # Train model
+    ent.build()
+    ent.run()
+
+    # Save output
+    ent.save(args.output)
+    print(f"Model saved to {args.output}")
+
+
+if __name__ == "__main__":
+    args = parse_args()
+    main(args)
+

### 2025-08-18 env.sh

@@ -1,0 +1,19 @@
+#!/usr/bin/env bash
+# Environment setup for fellowship project
+
+# Data path
+export data="/mnt/d/fellowship/m4efad/recovery/data"
+
+# Add project paths to PATH
+export PATH="code/:$PATH"
+export PATH="metatoolkit/metatoolkit/:$PATH"
+export PATH="maaslin3/R/:$PATH"
+export PATH="Maaslin2/R/:$PATH"
+
+# Activate Python virtual environment
+if [ -f venv/bin/activate ]; then
+    . venv/bin/activate
+else
+    echo "⚠️  No venv found at venv/bin/activate"
+fi
+

### 2025-08-21 code/mefisto_run.py
Removes the Brain Structure measurements

@@ -1,34 +1,63 @@
 #!/usr/bin/env python
-from mofapy2.run.entry_point import entry_point
 import pandas as pd
+from mofapy2.run.entry_point import entry_point
 
-# Load long-format input
-df = pd.read_csv("results/mofa/mefisto_input.tsv", sep='\t')
+# This part of the code is unchanged from your original script
+df = pd.read_csv("results/mofa/mefisto_input.tsv", sep="\t")
+df['sample'] = df['sample'].str.strip()
+df['group'] = df['group'].str.strip()
+df = df.set_index('sample', drop=False)
 
+# ----------------------------------------------------
+# Create samples_metadata
+# ----------------------------------------------------
+# We need to create a dictionary of dataframes, one for each group.
+# This structure is what the MOFA+ `set_covariates` method expects.
+samples_metadata_dict = {}
+
+# Iterate over each unique group in your main DataFrame
+for g in df['group'].unique():
+    # Filter the main DataFrame to get only the data for the current group
+    # We select the 'timepoint' and 'sample' columns
+    tmp_df = df[df['group'] == g][['timepoint', 'sample']]
+
+    # Drop any duplicate rows to ensure one row per sample in this group
+    tmp_df = tmp_df.drop_duplicates()
+
+    # Set the 'sample' column as the index
+    tmp_df = tmp_df.set_index('sample')
+
+    # Store this cleaned DataFrame in our dictionary with the group name as the key
+    samples_metadata_dict[g] = tmp_df
+
+# ----------------------------------------------------
+# Initialize and set options
+# ----------------------------------------------------
+# Initialize the entry point object
 ent = entry_point()
-ent.set_data_options(center_groups=False)
+
+# Set the main data DataFrame
 ent.set_data_df(df)
 
-# --- Create samples_metadata for each group ---
-groups = df['group'].unique()
-samples_metadata = []
-for g in groups:
-    # select unique rows for this group, with sampleID and covariates
-    tmp = df[df['group']==g][['sample', 'timepoint']].drop_duplicates().set_index('sample')
-    samples_metadata.append(tmp)
+# Now, set the covariates using the correctly structured dictionary
+# The keys of the dictionary ('g') will be recognized as the groups.
+# The `covariates_names` should be the name of the column you want to use as a covariate.
+ent.set_covariates(samples_metadata_dict, covariates_names="timepoint")
 
-ent.data_opts["samples_metadata"] = samples_metadata
-
[…]

### 2025-08-28 code/mefisto_extract.py
Minor edits

@@ -1,0 +1,27 @@
+#!/usr/bin/env python
+
+import mofax
+import argparse
+import pandas as pd
+from pathlib import Path
+
+parser = argparse.ArgumentParser(description="Extract weights and factors from mefisto model")
+parser.add_argument('--model')
+parser.add_argument('--output_dir')
+args = parser.parse_args()
+
+# Create output directory safely
+output_dir = Path(args.output_dir)
+output_dir.mkdir(parents=True, exist_ok=True)
+
+print("Loading MOFA model...")
+m = mofax.mofa_model(args.model)
+
+print('Extracting weights')
+weights = m.get_weights(df=True)
+weights.to_csv(output_dir / 'weights.tsv', sep='\t')
+
+print('Extracting factors')
+factors = m.get_factors(df=True)
+factors.to_csv(output_dir / 'factors.tsv', sep='\t')
+

### 2025-08-28 code/mefisto_plots.py

@@ -1,6 +1,6 @@
 #!/usr/bin/env python
 
-import mofax as mofa
+import mofax
 import argparse
 import matplotlib.pyplot as plt
 import seaborn as sns
@@ -8,14 +8,19 @@ import pandas as pd
 import numpy as np
 
 parser = argparse.ArgumentParser(description="Visualize R² from MOFA+ model")
-parser.add_argument('--model', default='results/mofa/t1.hdf5')
+parser.add_argument('--model')
+parser.add_argument('--output')
 args = parser.parse_args()
 
 breakpoint()
 print("Loading MOFA model...")
-model = mofa.mofa_model(args.model)
-r2_long_df = model.get_variance_explained()
+m = mofax.mofa_model(args.model)
+#model = mofa.mofa_model(args.model)
+fig = mofax.plot_r2(m, cmap='Blues')
+fig.savefig(args.output)
+#r2_long_df = model.get_variance_explained()
 
+'''
 # Barplot
 plt.figure(figsize=(4, 3))
 sns.barplot(data=r2_long_df, x='View', y='R2', estimator=np.sum, errorbar=None)
@@ -31,4 +36,4 @@ sns.heatmap(r2_df, cmap='viridis', vmax=15)
 plt.title('R² per View and Factor')
 plt.tight_layout()
 plt.show()
-
+'''

### 2025-08-28 code/mefisto_run.py

@@ -1,63 +1,81 @@
 #!/usr/bin/env python
 import pandas as pd
+import sys
 from mofapy2.run.entry_point import entry_point
 
-# This part of the code is unchanged from your original script
-df = pd.read_csv("results/mofa/mefisto_input.tsv", sep="\t")
+# ----------------------------------------------------
+# Load and preprocess the data
+# ----------------------------------------------------
+data = sys.argv[1]
+output = sys.argv[2]
+
+# Read input data
+df = pd.read_csv(data, sep="\t")
 df['sample'] = df['sample'].str.strip()
 df['group'] = df['group'].str.strip()
 df = df.set_index('sample', drop=False)
 
 # ----------------------------------------------------
-# Create samples_metadata
+# Create samples_metadata dictionary
 # ----------------------------------------------------
-# We need to create a dictionary of dataframes, one for each group.
-# This structure is what the MOFA+ `set_covariates` method expects.
+unique_samples_df = df[['group', 'timepoint', 'sample']].drop_duplicates().set_index('sample')
 samples_metadata_dict = {}
 
-# Iterate over each unique group in your main DataFrame
 for g in df['group'].unique():
-    # Filter the main DataFrame to get only the data for the current group
-    # We select the 'timepoint' and 'sample' columns
-    tmp_df = df[df['group'] == g][['timepoint', 'sample']]
+    group_samples = df[df['group'] == g]['sample'].unique()
+    tmp_df = (
+        df.loc[df['sample'].isin(group_samples), ['sample', 'timepoint']]
+        .drop_duplicates()
+        .set_index('sample')
+    )
+    tmp_df = tmp_df.loc[group_samples]
 
-    # Drop any duplicate rows to ensure one row per sample in this group
-    tmp_df = tmp_df.drop_duplicates()
+    if tmp_df.shape[0] != len(group_samples):
+        print(f"⚠ Mismatch in group {g}: data has {len(group_samples)} samples, meta has {tmp_df.shape[0]}")
 
-    # Set the 'sample' column as the index
-    tmp_df = tmp_df.set_index('sample')
-
-    # Store this cleaned DataFrame in our dictionary with the group name as the key
     samples_metadata_dict[g] = tmp_df
 
 # ----------------------------------------------------
-# Initialize and set options
+# Initialize and configure MEFISTO
 # ----------------------------------------------------
-# Initialize the entry point object
 ent = entry_point()
 
-# Set the main data DataFrame
-ent.set_data_df(df)
[…]

### 2025-08-29 code/calculate_recovery.py
Lots of manuscript updates

@@ -2,53 +2,48 @@
 # -*- coding: utf-8 -*-
 
 import pandas as pd
-import seaborn as sns
 import matplotlib.pyplot as plt
 
-df = pd.read_csv('results/timepoints/yr2/anthro.tsv', sep='\t', index_col=0)
+df = pd.read_csv('results/recovery/anthro_52.tsv', sep='\t', index_col=0)
 meta = pd.read_csv('results/filtered/meta.tsv', sep='\t', index_col=0)
+ 
+# Recovered at yr2
+yr2_recovered = df['WLZ_WHZ'].gt(-1)
 
-# Just recovered
-recovered = meta.loc[meta.Recovery == 'Recovered']
-sus_recovered = df.loc[recovered.index, 'WLZ_WHZ'].gt(-1)
-meta['Sus_recovered'] = sus_recovered
+# Define categories
+meta['Recovery_status'] = 'No recovery'  # default
 
-counts = meta.reset_index().groupby(['Feed','Recovery','Sus_recovered']).count()['subjectID'].reset_index()
+# Recovery (sustained recovery at yr2)
+meta.loc[(meta.Recovery == 'Recovered') & yr2_recovered, 'Recovery_status'] = 'Sustained recovery'
 
-# meta must already have: Feed, Recovery, Sus_recovered
-def classify_recovery(row):
-    if row['Recovery'] == 'Recovered' and row['Sus_recovered']:
-        return 'Sustained recovery'
-    elif row['Recovery'] == 'Recovered':
-        return 'Recovered'
-    else:
-        return 'No recovery'
+# Unsustained recovery (recovered before 15mo, but not at yr2)
+meta.loc[(meta.Recovery == 'Recovered') & (~yr2_recovered), 'Recovery_status'] = 'Unsustained recovery'
 
-meta['Recovery_status'] = meta.apply(classify_recovery, axis=1)
+# Delayed recovery (not recovered before 15mo, but recovered at yr2)
+meta.loc[(meta.Recovery != 'Recovered') & yr2_recovered, 'Recovery_status'] = 'Delayed recovery'
 
 # --- Count and percentage calculation ---
-status_order = ['Recovered', 'Sustained recovery', 'No recovery']
+status_order = ['No recovery', 'Unsustained recovery', 'Delayed recovery', 'Sustained recovery']
 colors = {
-    'Recovered': '#91bfdb',           # light blue
-    'Sustained recovery': '#0571b0',  # darker blue
[…]

### 2025-08-29 code/recovery_barplot.py

@@ -1,0 +1,81 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+
+import pandas as pd
+import matplotlib.pyplot as plt
+
+df = pd.read_csv('results/timepoints/yr2/anthro.tsv', sep='\t', index_col=0)
+meta = pd.read_csv('results/filtered/meta.tsv', sep='\t', index_col=0)
+ 
+# Just filter MAM
+meta = meta.loc[meta.Condition == 'MAM']
+df = df.loc[meta.index]
+
+# Recovered at yr2
+yr2_recovered = df['WLZ_WHZ'].gt(-1)
+
+# Define categories
+meta['Recovery_status'] = 'No recovery'  # default
+
+# Recovery (sustained recovery at yr2)
+meta.loc[(meta.Recovery == 'Recovered') & yr2_recovered, 'Recovery_status'] = 'Sustained recovery'
+
+# Unsustained recovery (recovered before 15mo, but not at yr2)
+meta.loc[(meta.Recovery == 'Recovered') & (~yr2_recovered), 'Recovery_status'] = 'Unsustained recovery'
+
+# Delayed recovery (not recovered before 15mo, but recovered at yr2)
+meta.loc[(meta.Recovery != 'Recovered') & yr2_recovered, 'Recovery_status'] = 'Delayed recovery'
+
+# --- Count and percentage calculation ---
+status_order = ['No recovery', 'Unsustained recovery', 'Delayed recovery', 'Sustained recovery']
+colors = {
+    'Sustained recovery': '#0571b0',
+    'No recovery': '#d73027',
+    'Unsustained recovery': '#fc8d59',
+    'Delayed recovery': '#2ca25f'
+}
+
+counts = (
+    meta.groupby(['Feed', 'Recovery_status'])
+    .size()
+    .reset_index(name='Count')
+    .pivot(index='Feed', columns='Recovery_status', values='Count')
+    .fillna(0)
+    .reindex(columns=status_order)
+)
+
+percentages = counts.div(counts.sum(axis=1), axis=0) * 100
+
+# --- Plotting ---
+fig, ax = plt.subplots(figsize=(3.2, 1.8))
+
+bottom = None
+for status in status_order:
+    ax.bar(
+        counts.index,
+        percentages[status],
+        bottom=bottom,
+        color=colors[status],
+        label=status
+    )
+
+    for i, (pct, cnt) in enumerate(zip(percentages[status], counts[status])):
+        if pct > 0:
[…]

### 2025-08-30 writing/manuscript/main.tex
Writing

@@
 \usepackage{booktabs}
+\usepackage{graphicx}
+\graphicspath{{figures/}}
 
-\title{
-  Understanding Recovery from Malnutrition in Early Life:\\
-  A Multi-Omic Analysis of the Infant Gut Microbiome, Plasma Lipidome, and Neurodevelopmental Outcomes
-}
+\title{Understanding Recovery from Malnutrition in Early Life: A Multi-Omic Analysis of the Infant Gut Microbiome, Plasma Lipidome, and Neurodevelopmental Outcomes}
 
-\author[1]{Authors}
+%\author[1]{Authors}
 %\author[1,*]{Theo Portlock}
@@
 %
-\affil[1]{Affiliations}
+%\affil[1]{Affiliations}
 %\affil[1]{The Liggins Institute, University of Auckland, NZ}
@@
 p = 0.9460).
-However, a key distinction emerged in time to recovery, which was shorter by approximately ten days in the ERUSF group (p = 0.0422, d = 0.46) (Table 1)
+However, a key distinction emerged in time to recovery, which was shorter by approximately ten days in the ERUSF group (p = 0.0422, d = 0.46) (Table 1).
+TABLE1 INCLUDED SEPARATELY
 
@@
 	\centering
-	\includegraphics[scale=1.2]{figures/recovery2.pdf}
+	\includegraphics[scale=1.2]{figures/recovery2}
 	\caption[Study design and changes in weight-for-length z-scores (WLZ) in children with moderate acute malnutrition (MAM) receiving enhanced or local refeeding]{
@@
 	}
-	\label{Figure1}
+	\label{fig:Figure1}
 \end{figure}
 
-\subsection{What influenced recovery}
+\subsection{What influenced recovery - Baseline anthropometrics, feeding failures, and genetic predisposition}
 % A background to the data collected
@@
 % Prediction of recovery from baseline facors - MOFA
+Prediction of recovery was done using MOFA to incorporate intermodal interactions.
 % risk factors for sustained recovery
@@
 % Why mixed models - maaslin3
+% Difficulties of interpretation of multiple datasets - including missing data
+% MOFA and factorisation
+% Machine learning pipeline explaination
+\begin{figure}[H]
+	\centering
+	\includegraphics[scale=0.6]{figures/ev}
+	\caption[Factors that associate with recovery]{
+		Explained variance as measured by PERMANOVA of beta diversities between samples within each dataset
+	}
+\label{fig:Figure2}
+\end{figure}
 
@@
 The most significant factors were anthropometric such as head circumference, indicating extent of malnutrition.
+This was shortly followed by Feeding failure rates, and then by genetic predisposition.
 % Influence of genetics on recovery
@@
 
-\subsection{What did recovery influence}
-% Maaslin3
-
-\subsection{Factors improved/deteriorated during the treatment}
+\subsection{Factors improved/deteriorated during the treatment disproportionately between refeed methods}
 % Definition of improvement
@@
 	\centering
-	\includegraphics[scale=0.6]{figures/newrecoveryfeed3.pdf}
+	\includegraphics[scale=0.6]{figures/newrecoveryfeed2}
 	\caption[Refeed methods improve/deteriorate bodily systems differently]{
@@
 	}
-\label{Figure2}
+\label{fig:Figure3}
 \end{figure}
-
-\subsection{Using a systems biology approach to understand recovery}
-% Difficulties of interpretation of multiple datasets - including missing data
-% MOFA and factorisation
-% Machine learning pipeline explaination
 
@@
 % Network stats - number of nodes, edges, directional nodes, interaction effects
-It was found that proline metabolism was correlated with EEG and cognitive outcomes, suggesting neuro-metabolic recovery pathways.
+It was found that proline metabolism was correlated with cognitive outcomes, suggesting neuro-metabolic recovery pathways.
 
@@
 \centering
-\includegraphics[scale=0.5]{figures/mixedmodelnetwork.pdf}
+\includegraphics[scale=0.5]{figures/mixedmodelnetwork}
 	\caption[Mechanism of E-RUTF differential impacts on brain structure]{
@@
 	}
-\label{Figure3}
+\label{Figure4}
 \end{figure}
 
+\subsection{Digital twin modelling to optimise novel refeed nutritional contents}
+
 \section{Discussion}
-
-\section{Conclusion}
 
@@
 Neurological measures, Bailey scores, EEG data were collected upon enrolment to evaluate neurological development.
-
-\subsection{EEG data collection and analysis}
-Continuous scalp EEG was recorded using NetStation 4.5.4. and 128-channel Hydrocel Geodesic Sensor Nets modified to remove eye electrodes (Electrical Geodesics, Inc. (EGI), Eugene, OR, USA).
-Data was sampled at 500 Hz.
-Impedances were kept under 100 k\textohm{} when possible and measured once at the beginning of the session, and again halfway through.
-Sessions were conducted in a dimly lit room with the participants sitting on the parent’s lap.
-The participants were separated from the research staff conducting the session by a curtain, but the testing area was not acoustically or electrically shielded.
-A second research staff member was present in the testing area to help keep the participant engaged.
-EEG sessions consisted of 6 paradigms, i.e., resting state, visual working memory, flanker, disengagement, visual evoked potential, and auditory stimuli.
-The subsequent (pre-)processing steps were applied to the resting state data where participants watched a 3-minute video that featured toys.
-
-EEG data were preprocessed offline with MatLab (R2021B) using the Harvard Automated Processing Pipeline for Electroencephalography (HAPPE) Version 3 (Gabard-Durnam et al., 2018).
-A specified subset of 30 channels was excluded (‘E1’, ’E8’, ’E14’, ’E17’, ’E21’, ‘E25’, ’E32’, ‘E38’, ‘E43’, ’E44’,’ E48’, ’E49’, ’E56’, ’E63’, ’E68’, ’E73’, ’E81’, ’E88’, ’E94’, ’E99’, ’E107’, ’E113’, ’E114’, ’E119’, ’E120’, ’E121’, ‘E125', 'E126', 'E127', 'E128').
-Data were downsampled to 250Hz, bandpass filtered (1-100Hz), and filtered using a 50Hz cleanline filter for line noise removal.
-Bad channels were then automatically identified and rejected, and wavelet-enhanced Independent Component Analysis (ICA) and the Multiple Artifact Rejection Algorithm (MARA) were performed to detect and impute artifacts.
-Resting state data were segmented into 2s epochs; epochs with an amplitude >±150mV were rejected.
-Segments were also rejected using segment similarity criteria.
-Data were then re-referenced to the average of all channels.
-
-EEG outputs from HAPPE were then reformatted and processed using the Batch Electroencephalography Automated Processing Platform (BEAPP) (Levin et al., 2018) to extract power spectra for each participant across the following frequency bands: delta (2-4Hz), theta (4-6Hz), low alpha (6-9Hz), high alpha (9-12Hz), beta (12-30Hz), and gamma (30-45Hz) and the following regions of interest (see Supp Figure 2): occipital (‘E70’, ’E71’, ’E75’, ‘E76’, ‘E83’), temporal (‘E36’,‘E40’, ‘E41’, ‘E45’, ‘E46’, ‘E102’, ‘E103’, ‘E104’, ‘E108’, ‘E109’), parietal (‘E52’, ‘E53’, ‘E59’, ‘E60’, ‘E85’, ‘E86’, ‘E91’, ‘E92’), and frontal (‘E5’, ‘E6’, ‘E12’, ‘E13’, ‘E24’, ‘E27’, ‘E28’, ‘E33’, ‘E34’, ‘E112’, ‘E116’, ‘E117’, ‘E122’, ‘E123’, ‘E124’).
-Further, PSD values were normalized by a Log\textsubscript{10} transform.
 
@@
 For full methodological details see the supplementary information.
+NEED VITAMIN AND AA METHODS FROM CHRIS.
 
@@
 \centering
-\includegraphics[scale=0.8]{figures/newalldata.pdf}
+\includegraphics[scale=0.8]{figures/newalldata}
 	\caption[Datasets recoreded in the M4EFaD study]{
@@
 	}
-\label{Figure1}
+\label{fig:FigureS1}
 \end{figure}

### 2025-09-01 code/mefisto_run.py
Re-adds the baseline mefo analysis

@@ -56,7 +56,7 @@ ent.set_data_options(scale_views=True)
 
 # Model options: Start with 20 factors, let dropping handle pruning
 ent.set_model_options(
-    factors=20,  # initial high number
+    factors=10,  # initial high number
 )
 
 # Train options with factor dropping enabled

### 2025-09-14 analysis.sh
Working with cats clusters
Also clearing out the maaslin2 stuff as well as updating predict.
Lots of updates with this one

@@ -1,173 +1,0 @@
-#!/bin/bash
-# This script is used to run the commands for the analysis pipeline.
-# It is assumed that the commands are run in the order they are listed.
-# The script is not intended to be run as a standalone script, but rather as a template for running the commands in the pipeline.
-
-export PATH="code/:$PATH"
-export PATH="metatoolkit/metatoolkit/:$PATH"
-
-# Setup
-format_bfpob.py
-format_lipids.py
-format_anthro.py
-format_fnirs.py
-format_pots.py
-format_bayley.py
-format_vitamin.py
-format_deliverysupp.py
-format_micro.py
-format_psd.py
-format_microbiome_samplesheet.py
-format_glitter.py
-format_head.py
-format_kneaddata.py
-format_humann.py
-format_aa.py
-format_vep.py
-format_pss.py
-format_surveillance.py
-format_genetics.py
-format_metaphlan.py
-format_wolkes.py
-format_ses.py
-format_sleep.py
-format_pci.py
-calculate_diversity.sh results/metaphlan.tsv
-extract_species.py
-format_feedsandrecovery.py
-format_fcis.py
-merge.py premeta feedandrecovery deliverysupp bfpob -j outer -o meta
-parallel -j1 remove_3yr_control.py results/{}.tsv -o results/filtered/{}.tsv < conf/all_datasets.txt
-create_mapping.py -d results/filtered/ -i conf/timedatasets.txt
-join.py mapping results/filtered/meta.tsv --on subjectID -o timemeta
-
-# Time to recovery figure
-multiline.py --subject results/anthro.tsv \
-	--x timepoint \
-	--y WLZ_WHZ \
-	--hue Recovery \
-	--id subjectID \
-	--meta results/timemeta.tsv \
-	--output results/anthroline.svg
-
-# Explained variance
-# === NOTIME: Explained variance on unstratified datasets ===
-parallel -j1 onehot.py results/filtered/{}.tsv -o results/notime/{}.tsv < conf/notimedatasets.txt
-drop.py --infile results/notime/surveillance.tsv --labels days_of_catchup --axis 1 --outfile results/notime/surveillance.tsv
-mv results/notime/meta.tsv results/meta_onehot.tsv
[…]

### 2025-09-14 code/cat_cluster.py

@@ -291,3 +291,11 @@ def create_full_clustermap(
 
 
 create_full_clustermap(df, meta, top_n_features=10, output="results/clustermap_with_metadata.svg")
+
+
+
+
+# For polar plot export
+odf = plotdf.groupby(['Cat_cluster', 'PRS']).mean()
+oodf = odf.unstack().xs('Score_z', axis=1)
+oodf.to_csv('results/polardata.tsv', sep='\t')

### 2025-09-16 writing/manuscript/main.tex
More on figure 2

@@
 
-%\author[1]{Authors}
-%\author[1,*]{Theo Portlock}
-%\author[2,*]{Talat Sharma}
-%\author[2,*]{Shahria Hafiz Kakon}
-%\author[3,*]{Berit Hartjen}
-%\author[1,*]{Chris Pook}
-%\author[1,*]{Brooke Wilson}
-%\author[3]{Ayisha Bhuttor}
-%\author[1]{Daniel Ho}
-%\author[1]{Inoli Shennon Wadumesthrige Don}
-%\author[3]{Anne-Michelle Engelstad}
-%\author[3]{Renata Di Lorenzo}
-%\author[3]{Garrett Greaves}
-%\author[3]{Caroline Kelsey}
-%\author[1]{Peter Gluckman}
-%\author[1,4,5,6]{Justin O'Sullivan}
-%\author[7]{Terrence Forrester}
-%\author[3]{Charles Nelson}
-%
-%\affil[1]{Affiliations}
-%\affil[1]{The Liggins Institute, University of Auckland, NZ}
-%\affil[2]{Infectious Diseases Division, International Centre for Diarrheal Disease Research, Bangladesh}
-%\affil[3]{Department of Pediatrics, Boston Children’s Hospital and Harvard Medical School; Harvard Graduate School of Education, Boston, USA}
-%\affil[4]{The Maurice Wilkins Centre, The University of Auckland, New Zealand}
-%\affil[5]{MRC Lifecourse Epidemiology Unit, University of Southampton, UK}
-%\affil[6]{Singapore Institute for Clinical Sciences, A*STAR, Singapore}
-%\affil[7]{Faculty of Medical Sciences, UWI Solutions for Developing Countries, The University of the West Indies, Jamaica}
-%\affil[*]{These authors contributed equally}
+\author[1]{Authors}
+\author[1,*]{Theo Portlock}
+\author[2,*]{Talat Sharma}
+\author[2,*]{Shahria Hafiz Kakon}
+\author[3,*]{Berit Hartjen}
+\author[1,*]{Chris Pook}
+\author[1,*]{Brooke Wilson}
+\author[3]{Ayisha Bhuttor}
+\author[1]{Daniel Ho}
+\author[1]{Inoli Shennon Wadumesthrige Don}
+\author[3]{Anne-Michelle Engelstad}
+\author[3]{Renata Di Lorenzo}
+\author[3]{Garrett Greaves}
+\author[3]{Caroline Kelsey}
+\author[1]{Peter Gluckman}
+\author[1,4,5,6]{Justin O'Sullivan}
+\author[7]{Terrence Forrester}
+\author[3]{Charles Nelson}
+
+\affil[1]{Affiliations}
+\affil[1]{The Liggins Institute, University of Auckland, NZ}
+\affil[2]{Infectious Diseases Division, International Centre for Diarrheal Disease Research, Bangladesh}
+\affil[3]{Department of Pediatrics, Boston Children’s Hospital and Harvard Medical School; Harvard Graduate School of Education, Boston, USA}
+\affil[4]{The Maurice Wilkins Centre, The University of Auckland, New Zealand}
+\affil[5]{MRC Lifecourse Epidemiology Unit, University of Southampton, UK}
+\affil[6]{Singapore Institute for Clinical Sciences, A*STAR, Singapore}
+\affil[7]{Faculty of Medical Sciences, UWI Solutions for Developing Countries, The University of the West Indies, Jamaica}
+\affil[*]{These authors contributed equally}
 
@@
 
-\section{Abstract}
+\section*{Abstract}
 Recovery from undernutrition during childhood has historically centered on short term weight gain with little attention given to redressing the delays in neurocognitive development.
@@
 
-\section{Main}
+\section*{Main}
 % Definition of MAM
@@
 
-\section{Results}
-\subsection{Study population characteristics}
+\section*{Results}
+\subsection*{Study population characteristics}
 % Why was Bangladesh picked
@@
 
-\subsection{What influenced recovery - Baseline anthropometrics, feeding failures, and genetic predisposition}
+\subsection*{What influenced recovery - Baseline anthropometrics, feeding failures, and genetic predisposition}
 % A background to the data collected
@@
 
-\subsection{Factors improved/deteriorated during the treatment disproportionately between refeed methods}
+\subsection*{Factors improved/deteriorated during the treatment disproportionately between refeed methods}
 % Definition of improvement
@@
 
-\subsection{Network-Based Integration of Multimodal Changes to understand recovery mechanism}
+\subsection*{Network-Based Integration of Multimodal Changes to understand recovery mechanism}
 % Networks as a tool to understand data relationships
@@
 
-\subsection{Digital twin modelling to optimise novel refeed nutritional contents}
+\subsection*{Digital twin modelling to optimise novel refeed nutritional contents}
 
-\section{Discussion}
+\section*{Discussion}
 
-\section{Methods}
+\section*{Methods}
 PSS method \cite{mozumder2022reliability}
 
-\subsection{Ethics}
+\subsection*{Ethics}
 The M4EFaD intervention was registered NCT05629624 on clinicaltrials.gov.
@@
 
-\subsection{Study Design and Participants}
+\subsection*{Study Design and Participants}
 The study was performed on the baseline, and two year data from three cohorts of infants who were enrolled (between Jan – December 2022) as part of the M4EFaD intervention within the Mirpur slum, Dhaka, Bangladesh.
@@
 
-\subsection{Recruitment and anthropometric data collection}
+\subsection*{Recruitment and anthropometric data collection}
 Enrolment was initiated on February 7, 2022, and continued until February 2024.
@@
 
-\subsection{Developmental Outcomes} 
+\subsection*{Developmental Outcomes} 
 The Bayley Scales of Infant and Toddler Development, Fourth Edition (BSID-IV) cognitive, language, and motor subscales were administered to all participants.
@@
 
-\subsection{Biological sample collection}
+\subsection*{Biological sample collection}
 Stool samples were collected from each infant at their home at the baseline visit.
@@
 
-\subsection{Microbiome DNA extraction and sequencing}
+\subsection*{Microbiome DNA extraction and sequencing}
 DNA was extracted from stool samples using the ZymoBIOMICS MagBead DNA/RNA extraction kit (Zymo Research, \#R2136) following the standard protocol.
@@
 
-\subsection{Plasma lipidomics}
+\subsection*{Plasma lipidomics}
 Plasma samples for lipidomics were thawed on ice and extracted according to a method modified from \citet{liu2016plasma}.
@@
 
-\subsection{Statistical Analyses}
+\subsection*{Statistical Analyses}
 Python version 3.9.2 was used to perform all analysis \cite{van1995python}.
@@
 
-\subsection{Machine learning}
+\subsection*{Machine learning}
 SHAP Value (SHapley Additive exPlanations) interpretation was used to interpret the contributions each feature had on the model's performance using the ‘shap’ python package \cite{lundberg2017unified}.
 
-\subsection{Network analysis}
+\subsection*{Network analysis}
 Absolute spearman rho of above 0.3 were used as edges and gut bacterial species and functional profiles, EEG, and plasma lipids were used as nodes coloured by their mean SHAP scores for classifier models that distinguish MAM from well-nourished conditions.
 
-\section{Code availability}
+\section*{Code availability}
 All analysis code is available on the GitHub repository.
@@
 
-\section{Ethics approval and consent to participate}
+\section*{Ethics approval and consent to participate}
 Ethical approvals were obtained from the Research Review Committee (RRC; August 21, 2021) and Ethical Review Committee (ERC) of icddr,b (protocol no: PR-21084; September 21, 2021), Institutional Review Board of Boston Children’s Hospital, USA (for analyses of neuropsychological assessments), University of Auckland, New Zealand (approval AH23922; for analyses of collected biological samples) and University of West Indies (CREC-MN.51, 21/22).
 
-\section{Data availability}
+\section*{Data availability}
 EEG and metadata are available from the authors, upon reasonable request that meets the ethics of the study.
 
-\section{Competing interests}
+\section*{Competing interests}
 The authors declare that they have no competing interests.
 
-\section{Funding}
+\section*{Funding}
 Work on this clinical trial is supported by Wellcome Leap (9942 Culver Blvd Unit 1277 Culver City, CA 90232-4167, United States; www.wellcomeleap.org) to PDG, JMO, TF and CAN as part of the 1kD Program.
@@
 
-\section{Author Contributions}
+\section*{Author Contributions}
 TP, KG and JOS drafted and co-wrote the manuscript.
@@
 
-\section{Acknowledgements}
+\section*{Acknowledgements}
 The authors would like to acknowledge the participants in Mirpur, Dhaka, Bangladesh for their contributions to this study.
@@
 
-\section{Supplementary material}
+\section*{Supplementary materials}
 \begin{figure}[!htb]

### 2025-09-16 code/importance_pie.py

@@ -1,0 +1,116 @@
+#!/usr/bin/env python3
+"""
+Create a nested donut (sunburst-like) plot of SHAP values.
+
+Usage:
+    python sunburst_matplotlib.py \
+        --input results/prediction/dataset_rf_shap/mean_abs_shap_test.tsv \
+        --output sunburst_shap_matplotlib.svg \
+        --threshold 0.01
+"""
+import argparse
+import pandas as pd
+import matplotlib.pyplot as plt
+from matplotlib.colors import to_rgba
+
+
+# -------------------------------------------------------------------
+# Utility
+# -------------------------------------------------------------------
+def lighten(color, alpha=0.4):
+    """Return a lighter / transparent version of a color."""
+    r, g, b, _ = to_rgba(color)
+    return (r, g, b, alpha)
+
+
+def plot_nested_donut(df: pd.DataFrame, output_file: str, threshold: float = 0.0):
+    """
+    Nested donut:
+    - Inner ring: categories
+    - Outer ring: all features (keep arc), but
+      features < threshold are drawn in white and have no label.
+    """
+    # Category totals
+    cat_total = df.groupby("category")["test_mean_abs_shap"].sum()
+    cat_order = cat_total.sort_values(ascending=False)
+
+    # Base colors for each category
+    base_colors = plt.rcParams["axes.prop_cycle"].by_key()["color"]
+    cat_colors = {cat: base_colors[i % len(base_colors)]
+                  for i, cat in enumerate(cat_order.index)}
+
+    inner_sizes, inner_labels, inner_colors = [], [], []
+    outer_sizes, outer_labels, outer_colors = [], [], []
+
+    for cat in cat_order.index:
+        # Inner ring (categories)
+        inner_sizes.append(cat_total[cat])
+        inner_labels.append(f"{cat}\n({cat_total[cat]/cat_total.sum()*100:.1f}%)")
+        inner_colors.append(cat_colors[cat])
+
+        # Outer ring: keep every feature, but mask those < threshold
+        sub = (df[df["category"] == cat]
+               .sort_values("test_mean_abs_shap", ascending=False))
+        for _, row in sub.iterrows():
+            outer_sizes.append(row["test_mean_abs_shap"])
+            if row["test_mean_abs_shap"] >= threshold:
+                outer_labels.append(row["feature"])
+                outer_colors.append(lighten(cat_colors[cat], 0.5))
+            else:
+                outer_labels.append("")      # no text
+                outer_colors.append("white") # white wedge
+
[…]

### 2025-09-16 code/predict.sh

@@ -6,9 +6,7 @@ set +e
 
 source env.sh
 
-#data=results/merged_dataset.tsv
-data=results/timepoints/notp/surveillance.tsv
-meta=results/timepoints/yr2/anthro.tsv
+data=results/merged_dataset.tsv
 output=results/prediction
 
 rm -r $output

### 2025-10-16 code/calculate_recovery.py
All descriptive tables

@@ -53,6 +53,11 @@ meta.loc[
 # --- Optional: Check category counts ---
 print(meta['Recovery_status'].value_counts())
 
+# Add any recovered
+meta.loc[:, 'Recovered'] = (meta.Recovery_status != 'No recovery')
+meta.loc[meta.Recovered == 1 , 'Recovered'] = 'Recovered'
+meta.loc[meta.Recovered == 0, 'Recovered'] = 'No recovery'
+
 # Save
-meta.Recovery_status.to_frame().to_csv('results/filtered/recovery_status.tsv', sep='\t')
+meta[['Recovered','Recovery_status']].to_csv('results/filtered/recovery_status.tsv', sep='\t')
 

### 2025-10-16 code/descriptive_stats.sh

@@ -3,6 +3,8 @@ source env.sh
 
 #input=results/table1/work/all_data_merged.tsv
 #output=results/table1/work/stats
+input=results/table2/work/data_merged.tsv \
+output=results/table2/work/MAM_recovery_stats
 input=$1
 output=$2
 
@@ -19,15 +21,18 @@ filter.py $input \
 
 corr.py \
 	$output/numbers.tsv \
+	--dropna \
 	-o $output/corr.tsv
 
 mannwhitneyu.py \
 	$output/numbers.tsv \
 	$output/categories.tsv \
+	--dropna \
 	-o $output/mwu.tsv
 
 fisher.py \
 	$output/categories.tsv \
+	--dropna \
 	-o $output/fisher.tsv
 
 merge.py \
@@ -38,12 +43,3 @@ merge.py \
 	--add-filename \
 	--filename-format base \
 	-o $output/merged_stats.tsv
-
-filter.py \
-	$output/merged_stats.tsv \
-	-q 'p_value < 0.05' \
-	-o $output/merged_stats_filtered.tsv
-
-create_network.py \
-	--edges $output/merged_stats_filtered.tsv \
-	--output $output/merged_stats_filtered_network.graphml

### 2025-10-16 code/maaslin3_baseline.sh

@@ -2,28 +2,304 @@
 set -euo pipefail
 source env.sh
 
-# Need to sort this for figure 2
 filter.py results/filtered/timemetaMAM_0_52tp_filled.tsv -q 'timepoint == 0' -o results/filtered/timemetaMAM_0tp_filled.tsv
-maaslin3.R results/filtered/head.tsv results/filtered/timemetaMAM_0tp_filled.tsv results/maaslin/head --formula "~ Feed + Recovery" --reference "Feed,Local RUSF (A);Recovery,No recovery" --min_abundance 0 --min_prevalence 0.0 --max_significance 0.1 --normalization NONE --transform NONE --warn_prevalence False --small_random_effects True --evaluate_only abundance --max_pngs 100 --cores 1
 
-maaslin3.R results/filtered/economics.tsv results/filtered/meta_MAM.tsv results/maaslin/economics --formula "~ Feed + Recovery" --reference "Feed,Local RUSF (A);Recovery,No recovery" --min_abundance 0 --min_prevalence 0.0 --max_significance 0.1 --normalization NONE --transform NONE --warn_prevalence False --evaluate_only abundance --max_pngs 100 --cores 1
-maaslin3.R results/filtered/education.tsv results/filtered/meta_MAM.tsv results/maaslin/education --formula "~ Feed + Recovery" --reference "Feed,Local RUSF (A);Recovery,No recovery" --min_abundance 0 --min_prevalence 0.0 --max_significance 0.1 --normalization NONE --transform NONE --warn_prevalence False --evaluate_only abundance --max_pngs 100 --cores 1
-maaslin3.R results/filtered/family.tsv results/filtered/meta_MAM.tsv results/maaslin/family --formula "~ Feed + Recovery" --reference "Feed,Local RUSF (A);Recovery,No recovery" --min_abundance 0 --min_prevalence 0.0 --max_significance 0.1 --normalization NONE --transform NONE --warn_prevalence False --evaluate_only abundance --max_pngs 100 --cores 1
[…]

### 2025-10-16 code/maaslin3_healthy.sh

@@ -1,40 +1,234 @@
 #!/usr/bin/env bash
-source env.sh
 set -euo pipefail
+source env.sh
 
-filter.py results/filtered/timemeta.tsv -q 'timepoint == 0 or timepoint == 52' -o results/filtered/timemeta_0_52tp.tsv
-fillna.py -i results/filtered/timemeta_0_52tp.tsv -c Feed,Recovery -v 'Well-nourished' -o results/filtered/timemeta_0_52tp_filled.tsv
+filter.py \
+    results/filtered/timemeta.tsv \
+    -q 'timepoint == 0 or timepoint == 52' \
+    -o results/filtered/timemeta_0_52tp.tsv
+
+fillna.py \
+    -i results/filtered/timemeta_0_52tp.tsv \
+    -c Feed,Recovery \
+    -v 'Well-nourished' \
+    -o results/filtered/timemeta_0_52tp_filled.tsv
 
 # Test first here
-maaslin3.R results/filtered/species.tsv results/filtered/timemeta_0_52tp_filled.tsv results/maaslin/healthy/species --formula "~ timepoint*(Feed + Recovery) + (1|subjectID)" --reference "Feed,Well-nourished;Recovery,Well-nourished;timepoint,yr1" --min_abundance 0 --min_prevalence 0.2 --max_significance 0.1 --normalization TSS --transform LOG --warn_prevalence False --small_random_effects True --max_pngs 100 --cores 1
+maaslin3.R \
+    results/filtered/species.tsv \
+    results/filtered/timemeta_0_52tp_filled.tsv \
+    results/maaslin/time/species \
+    --formula "~ timepoint*(Feed + Recovery) + (1|subjectID)" \
+    --reference "Feed,Well-nourished;Recovery,Well-nourished;timepoint,yr1" \
+    --min_abundance 0 \
+    --min_prevalence 0.2 \
+    --max_significance 0.1 \
+    --normalization TSS \
+    --transform LOG \
+    --warn_prevalence False \
+    --small_random_effects True \
+    --max_pngs 100 \
+    --cores 1
 
-pointplot.py \
-  results/filtered/bayley.tsv \
-  --meta results/filtered/timemeta_0_52tp_filled.tsv \
-  -x timepoint \
-  -y 'expressive_communication_score' \
-  --hue 'Recovery' \
-  --id subjectID \
-  -o results/testpoint2.svg
+maaslin3.R \
[…]

### 2025-10-16 code/plot_recipe.py

@@ -80,3 +80,6 @@ sns.clustermap(fdf, cmap='coolwarm', figsize=(1.37,4), yticklabels=True, row_clu
 plt.savefig(f'{output}/{var}.svg')
 plt.show()
 
+# Save
+df.to_csv('results/recipe.tsv')
+

### 2025-10-16 code/table2_refeed_recovery.sh

@@ -1,0 +1,89 @@
+#!/bin/bash
+
+source env.sh
+
+rm -rf results/table2
+mkdir -p results/table2/work/{selected,formatted,descriptive}
+
+onehot.py \
+	results/filtered/recovery_status.tsv \
+	--prefix-sep ', ' \
+	--dtype bool \
+	-o results/table2/work/formatted/recovery_status_oh.tsv
+
+select.py \
+	results/filtered/surveillance.tsv \
+	-c 'days_of_catchup,fail_no.failure' \
+	-o results/table2/work/formatted/surveillance.tsv
+
+select.py \
+	results/filtered/meta_MAM.tsv \
+	-c 'Feed' \
+	-o results/table2/work/selected/meta_MAM.tsv
+
+onehot.py \
+	results/table2/work/selected/meta_MAM.tsv \
+	--prefix-sep ', ' \
+	--dtype bool \
+	-o results/table2/work/formatted/meta_MAM_oh.tsv
+
+for file in results/table2/work/formatted/{meta_MAM_oh,recovery_status_oh,surveillance}.tsv; do
+    descriptive.py \
+        -i "$file" \
+        --meta results/filtered/meta.tsv \
+        -g 'Feed' \
+        -o "results/table2/work/descriptive/$(basename "$file")"
+done
+
+merge.py \
+	results/table2/work/descriptive/recovery_status_oh.tsv \
+	results/table2/work/descriptive/surveillance_fixed.tsv \
+	-a \
+	--add-filename \
+	--filename-format base \
+	-o results/table2/work/merged.tsv
+
+# Stats
+merge.py \
+	results/table2/work/formatted/recovery_status_oh.tsv \
+	results/table2/work/formatted/meta_MAM_oh.tsv \
+	results/table2/work/formatted/surveillance.tsv \
+	-o results/table2/work/data_merged.tsv
+
+descriptive_stats.sh \
+	results/table2/work/data_merged.tsv \
+	results/table2/work/MAM_recovery_stats
+
+filter.py \
+	results/table2/work/MAM_recovery_stats/merged_stats.tsv \
+	-q 'source == "Feed, ERUSF (B)"' \
+	--drop \
+	-o results/table2/work/MAM_recovery_stats/merged_stats_Recoverstats.tsv
+	
+select.py \
+	results/table2/work/MAM_recovery_stats/merged_stats_Recoverstats.tsv \
+	-c 'target,p_value' \
+	--drop-index \
[…]

### 2025-10-16 code/table3_recovery.sh

@@ -1,0 +1,59 @@
+#!/bin/bash
+
+source env.sh
+
+rm -rf results/table3
+mkdir -p results/table3/work/{selected,formatted,descriptive}
+
+select.py \
+	results/filtered/recovery_status.tsv \
+	-c Recovered \
+	-o results/table3/work/recovery.tsv
+
+onehot.py \
+	results/table3/work/recovery.tsv \
+	--dtype bool \
+	-o results/table3/work/formatted/recovery.tsv
+
+for file in results/table1/work/formatted/{Demographics,Parental_Education_and_Economics,Family_Structure,Household_characteristics,Other}.tsv; do
+    descriptive.py \
+        -i "$file" \
+        --meta results/table3/work/recovery.tsv \
+        -g 'Recovered' \
+        -o "results/table3/work/descriptive/$(basename "$file")"
+done
+
+merge.py \
+	results/table3/work/descriptive/* \
+	-a \
+	--add-filename \
+	--filename-format base \
+	-o results/table3/work/merged.tsv
+
+# For the stats
+merge.py \
+	results/table1/work/MAM_data_merged.tsv \
+	results/table3/work/formatted/recovery.tsv \
+	-o results/table3/work/MAM_data_merged_recovery.tsv
+
+descriptive_stats.sh \
+	results/table3/work/MAM_data_merged_recovery.tsv \
+	results/table3/work/MAM_recovery_stats
+
+filter.py \
+	results/table3/work/MAM_recovery_stats/merged_stats.tsv \
+	-q 'source == "Recovered.Recovered"' \
+	--drop \
+	-o results/table3/work/MAM_recovery_stats/merged_stats_Recoverstats.tsv
+	
+select.py \
+	results/table3/work/MAM_recovery_stats/merged_stats_Recoverstats.tsv \
+	-c 'target,p_value' \
+	--drop-index \
+	-o results/table3/work/MAM_recovery_stats/merged_stats_Recoverstats_vals.tsv
+
+# Merge
+merge.py \
+	results/table3/work/merged.tsv \
+	results/table3/work/MAM_recovery_stats/merged_stats_Recoverstats_vals.tsv \
+	-o results/table3/work/merged_stats.tsv
